---
title: "Advanced StyleRows and Conditional Formatting"
author: "ksTFL Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Advanced StyleRows and Conditional Formatting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette explains the powerful conditional row styling system in ksTFL, 
centered around `compute_cols()` and its action functions: `c_style()`, `c_merge()`, 
and `c_addrow()`. These tools enable sophisticated data-driven formatting without 
manual post-processing.

## Core Concept: Lazy Evaluation

`compute_cols()` uses **lazy evaluation**—it captures your conditions and actions 
as unevaluated expressions (quosures), storing them in the spec's metadata. The 
actual evaluation happens later during `create_report()`, when all specs are 
finalized and the data context is fully established.

### Why Lazy Evaluation?

1. **Deferred Context**: Column selections and data references are resolved when 
   the report structure is complete
2. **Tidyselect Support**: Use `everything()`, `starts_with()`, etc. without 
   worrying about timing
3. **Multi-Stage Processing**: Separate specification (design time) from execution 
   (render time)

## The Three Action Functions

### `c_style()`: Conditional Styling

Apply style references to cells based on conditions:

```{r c-style-basic}
library(ksTFL)

data <- data.frame(
  patient = sprintf("PAT-%03d", 1:20),
  age = c(23, 45, 67, 34, 89, 56, 42, 71, 38, 29,
          55, 66, 44, 52, 60, 48, 35, 70, 41, 58),
  response = sample(c("CR", "PR", "SD", "PD"), 20, replace = TRUE)
)

# Define styles first
spec <- create_table(data) |>
  add_style("highlight_green", s_font(color = "#006400", bold = TRUE)) |>
  add_style("highlight_red", s_font(color = "#8B0000", bold = TRUE))

# Apply conditional styling
spec <- spec |>
  compute_cols(
    response == "CR",
    c_style(response, styleRef = "highlight_green")
  ) |>
  compute_cols(
    response == "PD",
    c_style(response, styleRef = "highlight_red")
  )
```

**Result:** Rows where `response == "CR"` show green bold text; rows where 
`response == "PD"` show red bold text.

### `c_merge()`: Conditional Cell Merging

Merge multiple columns into a single display cell:

```{r c-merge-basic}
data_groups <- data.frame(
  group = c("Treatment A", "Treatment A", "Treatment A",
            "Placebo", "Placebo"),
  visit = c("Week 0", "Week 4", "Week 8", "Week 0", "Week 4"),
  value = c(5.2, 6.1, 7.3, 4.8, 5.5)
)

spec <- create_table(data_groups) |>
  compute_cols(
    group == lag(group, default = ""),  # Same group as previous row
    c_merge(c(group, visit), display_col = "group")
  )
```

**Result:** When consecutive rows have the same `group`, their `group` and `visit` 
cells merge, displaying the merged content in the `group` column.

### `c_addrow()`: Conditional Row Insertion

Insert new rows based on data patterns:

```{r c-addrow-basic}
spec <- create_table(data_groups) |>
  compute_cols(
    group != lag(group, default = ""),  # Group changes
    c_addrow(
      position = "before",
      values = list(
        group = paste0("=== ", group, " ==="),
        visit = "",
        value = NA
      )
    )
  )
```

**Result:** A summary row is inserted before each new group, with formatted 
group header text.

## Evaluation Context

### The Data Environment

Conditions and actions are evaluated in `spec$.metadata$data_env`, which contains:

1. **Data columns**: Direct access to all columns (even invisible ones)
2. **Helper functions**: `lag()`, `lead()`, `row_number()`, etc.
3. **Embedded functions**: Internal utilities for column access

```{r data-env-example}
# You can reference any column directly in conditions
spec <- create_table(data) |>
  compute_cols(
    age > 60 & response %in% c("CR", "PR"),  # Multi-column condition
    c_style(c(age, response), styleRef = "elderly_responder")
  )
```

### Available Helper Functions

ksTFL provides dplyr-like helpers evaluated in the data context:

- **`lag(x, n = 1, default = NA)`**: Previous row value(s)
- **`lead(x, n = 1, default = NA)`**: Next row value(s)
- **`row_number()`**: Current row index (1-based)
- **`n()`**: Total row count

```{r helpers-example}
# Highlight first and last rows
spec <- create_table(data) |>
  compute_cols(
    row_number() == 1 | row_number() == n(),
    c_style(everything(), styleRef = "border_emphasis")
  )

# Compare with previous row
spec <- spec |>
  compute_cols(
    age > lag(age, default = 0),
    c_style(age, styleRef = "increasing_trend")
  )
```

## Advanced `c_style()` Patterns

### Multiple Column Styling

Use tidyselect to style multiple columns at once:

```{r c-style-multi}
data_lab <- data.frame(
  patient = sprintf("PAT-%03d", 1:10),
  hemoglobin = rnorm(10, 13.5, 1.5),
  glucose = rnorm(10, 95, 15),
  cholesterol = rnorm(10, 200, 30)
)

spec <- create_table(data_lab) |>
  add_style("out_of_range", s_font(color = "#FF4500", bold = TRUE)) |>
  compute_cols(
    hemoglobin < 12 | hemoglobin > 16,
    c_style(hemoglobin, styleRef = "out_of_range")
  ) |>
  compute_cols(
    glucose < 70 | glucose > 140,
    c_style(glucose, styleRef = "out_of_range")
  ) |>
  compute_cols(
    cholesterol > 240,
    c_style(cholesterol, styleRef = "out_of_range")
  )
```

### Conditional Styling with Complex Logic

Combine multiple conditions:

```{r c-style-complex}
spec <- create_table(data) |>
  add_style("critical_senior", 
            s_font(color = "#8B0000", bold = TRUE),
            s_table_style(background_color = "#FFEBCD")) |>
  compute_cols(
    age >= 70 & response == "PD",
    c_style(c(patient, age, response), styleRef = "critical_senior")
  )
```

### Row-Level Styling

Style entire rows by targeting all columns:

```{r c-style-row}
spec <- create_table(data) |>
  add_style("alternate_row", 
            s_table_style(background_color = "#F0F0F0")) |>
  compute_cols(
    row_number() %% 2 == 0,  # Even rows
    c_style(everything(), styleRef = "alternate_row")
  )
```

## Advanced `c_merge()` Patterns

### Multi-Column Grouping Merges

Merge across multiple grouping levels:

```{r c-merge-multi}
data_nested <- data.frame(
  study = rep(c("Study A", "Study B"), each = 6),
  phase = rep(c("Phase I", "Phase II", "Phase III"), 4),
  site = rep(c("Site 1", "Site 2", "Site 1", "Site 2"), 3),
  enrollment = sample(10:50, 12)
)

spec <- create_table(data_nested) |>
  # Merge study column for consecutive same-study rows
  compute_cols(
    study == lag(study, default = ""),
    c_merge(study, display_col = "study")
  ) |>
  # Merge phase column within same study
  compute_cols(
    study == lag(study, default = "") & phase == lag(phase, default = ""),
    c_merge(phase, display_col = "phase")
  )
```

### Conditional Display Column Content

Customize merged cell content:

```{r c-merge-custom}
spec <- create_table(data_groups) |>
  compute_cols(
    group == lag(group, default = ""),
    c_merge(
      c(group, visit),
      display_col = "group",
      sep = " | "  # Custom separator for merged values
    )
  )
```

### Merging with Styling

Combine merging with conditional styles:

```{r c-merge-style}
spec <- create_table(data_nested) |>
  add_style("merged_header", 
            s_font(bold = TRUE),
            s_table_style(background_color = "#E0E0E0")) |>
  compute_cols(
    study == lag(study, default = ""),
    c_merge(study, display_col = "study")
  ) |>
  compute_cols(
    study != lag(study, default = ""),  # First row of group
    c_style(study, styleRef = "merged_header")
  )
```

## Advanced `c_addrow()` Patterns

### Summary Rows

Insert calculated summary rows:

```{r c-addrow-summary}
data_sales <- data.frame(
  region = c("North", "North", "South", "South", "West", "West"),
  product = rep(c("A", "B"), 3),
  revenue = c(100, 150, 200, 120, 180, 160)
)

spec <- create_table(data_sales) |>
  add_style("summary_row", 
            s_font(bold = TRUE),
            s_table_style(background_color = "#D3D3D3")) |>
  compute_cols(
    region != lead(region, default = ""),  # Last row of group
    c_addrow(
      position = "after",
      values = list(
        region = paste0(region, " Total"),
        product = "",
        revenue = sum(revenue[region == region])  # Will be calculated at eval time
      ),
      styleRef = "summary_row"
    )
  )
```

### Header Rows

Insert section headers:

```{r c-addrow-header}
spec <- create_table(data_sales) |>
  add_style("section_header",
            s_font(bold = TRUE, font_size = 12),
            s_table_style(background_color = "#4682B4")) |>
  compute_cols(
    region != lag(region, default = ""),  # First row of new region
    c_addrow(
      position = "before",
      values = list(
        region = toupper(region),
        product = "REGION HEADER",
        revenue = NA
      ),
      styleRef = "section_header"
    )
  )
```

### Conditional Row Insertion Logic

Insert rows only when specific conditions are met:

```{r c-addrow-conditional}
spec <- create_table(data_sales) |>
  compute_cols(
    revenue > 150 & product == "A",  # Only for high-revenue product A
    c_addrow(
      position = "after",
      values = list(
        region = paste0("⚠ ", region, " - High Performance"),
        product = "",
        revenue = revenue * 1.1  # Projected next period
      )
    )
  )
```

## Combining Actions

### Sequential Actions

Chain multiple `compute_cols()` calls for layered effects:

```{r combine-sequential}
spec <- create_table(data_sales) |>
  # Step 1: Add summary rows
  compute_cols(
    region != lead(region, default = ""),
    c_addrow(position = "after", 
             values = list(region = paste(region, "Total"), 
                          product = "", 
                          revenue = NA))
  ) |>
  # Step 2: Style summary rows
  add_style("bold_summary", s_font(bold = TRUE)) |>
  compute_cols(
    grepl("Total$", region),
    c_style(region, styleRef = "bold_summary")
  ) |>
  # Step 3: Merge region cells
  compute_cols(
    region == lag(region, default = ""),
    c_merge(region, display_col = "region")
  )
```

### Multiple Actions in One `compute_cols()`

**Note:** Currently, `compute_cols()` accepts one action function per call. To 
apply multiple actions to the same condition, use separate calls with identical 
conditions:

```{r combine-same-condition}
spec <- create_table(data) |>
  # Action 1: Style cells
  compute_cols(
    age > 65,
    c_style(age, styleRef = "elderly")
  ) |>
  # Action 2: Merge cells (same condition)
  compute_cols(
    age > 65,
    c_merge(c(patient, age), display_col = "patient")
  )
```

## Performance Tips

### 1. Minimize `compute_cols()` Calls

Each call adds evaluation overhead. Combine conditions when possible:

```{r perf-minimize}
# ❌ Less efficient:
spec <- create_table(data) |>
  compute_cols(age < 30, c_style(age, styleRef = "young")) |>
  compute_cols(age >= 30 & age < 60, c_style(age, styleRef = "middle")) |>
  compute_cols(age >= 60, c_style(age, styleRef = "senior"))

# ✅ More efficient (single pass with case_when logic):
spec <- create_table(data) |>
  add_style("young", s_font(color = "#008000")) |>
  add_style("middle", s_font(color = "#0000FF")) |>
  add_style("senior", s_font(color = "#FF0000")) |>
  compute_cols(
    age < 30,
    c_style(age, styleRef = "young")
  ) |>
  compute_cols(
    age >= 30 & age < 60,
    c_style(age, styleRef = "middle")
  ) |>
  compute_cols(
    age >= 60,
    c_style(age, styleRef = "senior")
  )
  
# Note: Consider refactoring to reduce compute_cols() calls if performance is critical
```

### 2. Use Vectorized Conditions

Avoid row-by-row operations in custom functions:

```{r perf-vectorized}
# ❌ Slower (scalar logic):
spec <- create_table(data) |>
  compute_cols(
    sapply(response, function(x) x %in% c("CR", "PR")),  # Row-by-row
    c_style(response, styleRef = "responder")
  )

# ✅ Faster (vectorized):
spec <- create_table(data) |>
  compute_cols(
    response %in% c("CR", "PR"),  # Vectorized
    c_style(response, styleRef = "responder")
  )
```

### 3. Pre-Filter Data When Possible

If conditions apply to a small subset, consider filtering data upfront:

```{r perf-prefilter}
# If only 5% of rows need special formatting:
# Consider creating separate tables and combining in report

data_outliers <- subset(data, age > 80)
data_normal <- subset(data, age <= 80)

spec_outliers <- create_table(data_outliers) |>
  add_style("outlier", s_font(color = "#FF0000", bold = TRUE)) |>
  compute_cols(TRUE, c_style(everything(), styleRef = "outlier"))

spec_normal <- create_table(data_normal)

# Combine in report
report <- create_report(spec_normal, spec_outliers)
```

### 4. Style Consolidation

ksTFL automatically consolidates identical styles, but you can help by reusing 
style references:

```{r perf-consolidation}
# ✅ Define once, use many times:
spec <- create_table(data) |>
  add_style("critical", s_font(color = "#FF0000", bold = TRUE)) |>
  compute_cols(age > 80, c_style(age, styleRef = "critical")) |>
  compute_cols(response == "PD", c_style(response, styleRef = "critical"))

# ❌ Avoid duplicate style definitions:
# (This creates two identical but separate styles)
spec <- create_table(data) |>
  add_style("critical_age", s_font(color = "#FF0000", bold = TRUE)) |>
  add_style("critical_response", s_font(color = "#FF0000", bold = TRUE))
```

## Debugging and Inspection

### Viewing Captured Actions

Inspect what `compute_cols()` has stored:

```{r debug-inspect}
spec <- create_table(data) |>
  compute_cols(
    age > 60,
    c_style(age, styleRef = "elderly")
  )

# Examine metadata
str(spec$.metadata$styleRows)
# Shows captured quosures and action types
```

### Testing Conditions Manually

Extract data and test conditions before adding to spec:

```{r debug-test}
# Get data from spec
data_test <- spec$.metadata$data_env$`__data__`

# Test your condition
test_condition <- with(data_test, age > 60)
sum(test_condition)  # How many rows match?

# Verify subset
data_test[test_condition, ]
```

### Incremental Building

Add `compute_cols()` one at a time and inspect results:

```{r debug-incremental}
spec <- create_table(data)

# Add first action
spec <- spec |>
  compute_cols(age > 60, c_style(age, styleRef = "elderly"))
print(spec)  # Check structure

# Add second action
spec <- spec |>
  compute_cols(response == "CR", c_style(response, styleRef = "success"))
print(spec)  # Check again
```

## Common Patterns Library

### Pattern 1: Alternating Row Colors

```{r pattern-alternate}
spec <- create_table(data) |>
  add_style("gray_bg", s_table_style(background_color = "#F5F5F5")) |>
  compute_cols(
    row_number() %% 2 == 0,
    c_style(everything(), styleRef = "gray_bg")
  )
```

### Pattern 2: Grouped Section Headers with Merging

```{r pattern-grouped}
spec <- create_table(data_groups) |>
  add_style("group_header", 
            s_font(bold = TRUE, font_size = 11),
            s_table_style(background_color = "#D0D0D0")) |>
  # Insert header row before each new group
  compute_cols(
    group != lag(group, default = ""),
    c_addrow(
      position = "before",
      values = list(group = paste("===", group, "==="), visit = "", value = NA),
      styleRef = "group_header"
    )
  ) |>
  # Merge consecutive same-group cells
  compute_cols(
    group == lag(group, default = ""),
    c_merge(group, display_col = "group")
  )
```

### Pattern 3: Conditional Highlighting with Thresholds

```{r pattern-thresholds}
spec <- create_table(data_lab) |>
  add_style("low", s_font(color = "#0000FF")) |>
  add_style("normal", s_font(color = "#008000")) |>
  add_style("high", s_font(color = "#FF0000")) |>
  compute_cols(
    hemoglobin < 12,
    c_style(hemoglobin, styleRef = "low")
  ) |>
  compute_cols(
    hemoglobin >= 12 & hemoglobin <= 16,
    c_style(hemoglobin, styleRef = "normal")
  ) |>
  compute_cols(
    hemoglobin > 16,
    c_style(hemoglobin, styleRef = "high")
  )
```

### Pattern 4: Comparative Analysis (vs. Previous Row)

```{r pattern-comparative}
spec <- create_table(data) |>
  add_style("improved", s_font(color = "#006400")) |>
  add_style("worsened", s_font(color = "#8B0000")) |>
  compute_cols(
    !is.na(lag(age)) & age > lag(age),
    c_style(age, styleRef = "improved")
  ) |>
  compute_cols(
    !is.na(lag(age)) & age < lag(age),
    c_style(age, styleRef = "worsened")
  )
```

### Pattern 5: Summary Rows with Totals

```{r pattern-totals}
# Note: This pattern requires careful handling of aggregates
# Currently, you can reference group-level calculations in values list

spec <- create_table(data_sales) |>
  add_style("total_row", 
            s_font(bold = TRUE),
            s_table_style(background_color = "#FFD700")) |>
  compute_cols(
    region != lead(region, default = ""),
    c_addrow(
      position = "after",
      values = list(
        region = "SUBTOTAL",
        product = "",
        revenue = NA  # Will be calculated during evaluation
      ),
      styleRef = "total_row"
    )
  )
```

## Limitations and Workarounds

### Limitation 1: No Direct Aggregate Functions in Conditions

You can't use `sum()`, `mean()`, etc. directly in conditions:

```{r limit-aggregates, error=TRUE}
# ❌ This won't work as expected:
spec <- create_table(data) |>
  compute_cols(
    age > mean(age),  # Evaluates mean() at condition capture, not evaluation
    c_style(age, styleRef = "above_average")
  )
```

**Workaround:** Pre-calculate and add as a column:

```{r workaround-aggregates}
data$age_above_avg <- data$age > mean(data$age)

spec <- create_table(data) |>
  compute_cols(
    age_above_avg,
    c_style(age, styleRef = "above_average")
  )
```

### Limitation 2: No Nested `c_*()` Functions

You can't nest action functions:

```{r limit-nested, error=TRUE}
# ❌ This is invalid:
spec <- create_table(data) |>
  compute_cols(
    age > 60,
    c_style(age, styleRef = c_merge(patient, age))  # Not allowed
  )
```

**Workaround:** Use separate `compute_cols()` calls:

```{r workaround-nested}
spec <- create_table(data) |>
  compute_cols(age > 60, c_style(age, styleRef = "elderly")) |>
  compute_cols(age > 60, c_merge(c(patient, age), display_col = "patient"))
```

### Limitation 3: Style References Must Exist

`styleRef` must reference a previously defined style:

```{r limit-styleref, error=TRUE}
# ❌ This will error at evaluation time:
spec <- create_table(data) |>
  compute_cols(age > 60, c_style(age, styleRef = "undefined_style"))
```

**Workaround:** Always define styles before using them:

```{r workaround-styleref}
spec <- create_table(data) |>
  add_style("elderly", s_font(bold = TRUE)) |>  # Define first
  compute_cols(age > 60, c_style(age, styleRef = "elderly"))
```

## Integration with Other ksTFL Features

### Using with `define_cols()`

`compute_cols()` works alongside column definitions:

```{r integrate-define}
spec <- create_table(data) |>
  define_cols(age, type = "numeric", format = "0.0", colWidth = "15%") |>
  compute_cols(
    age > 65,
    c_style(age, styleRef = "elderly")
  )
```

### Using with Invisible Columns

Reference invisible columns in conditions:

```{r integrate-invisible}
data$flag <- sample(c(TRUE, FALSE), nrow(data), replace = TRUE)

spec <- create_table(data) |>
  define_cols(flag, isVisible = FALSE) |>  # Hide column
  compute_cols(
    flag == TRUE,  # But use it in condition
    c_style(response, styleRef = "flagged")
  )
```

### Using in Multi-Spec Reports

Each spec can have independent `compute_cols()` logic:

```{r integrate-report}
spec1 <- create_table(data[1:10, ]) |>
  compute_cols(age > 60, c_style(age, styleRef = "elderly"))

spec2 <- create_table(data[11:20, ]) |>
  compute_cols(response == "CR", c_style(response, styleRef = "success"))

report <- create_report(spec1, spec2)
```

## Best Practices

1. **Define styles first**: Use `add_style()` before `compute_cols()`
2. **Test conditions incrementally**: Add one `compute_cols()` at a time
3. **Use meaningful style names**: "elderly" is clearer than "style1"
4. **Document complex logic**: Add comments explaining condition rationale
5. **Prefer vectorized operations**: Avoid row-by-row functions where possible
6. **Reuse styles**: Define once, reference many times for consistency
7. **Keep conditions simple**: Complex logic is harder to debug
8. **Check for NA values**: Use `!is.na()` when comparing with `lag()`/`lead()`

## Summary

- **`compute_cols()`**: Captures conditions and actions via lazy evaluation
- **`c_style()`**: Apply conditional styling to cells
- **`c_merge()`**: Merge cells across columns based on conditions
- **`c_addrow()`**: Insert new rows dynamically
- **Evaluation context**: Data environment with helper functions (`lag`, `lead`, `row_number`, `n`)
- **Performance**: Minimize calls, use vectorized operations, consolidate styles
- **Debugging**: Inspect metadata, test conditions manually, build incrementally

For styling fundamentals, see `vignette("Styling_Guide_with_ksTFL")`.  
For complete examples, see `vignette("Reporting_Examples_with_ksTFL")`.
