---
title: "Getting Started with ksTFL"
author: "ksTFL Team"
output:
  rmarkdown::html_vignette:
    dev: pdf
vignette: >
  %\VignetteIndexEntry{Getting Started with ksTFL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "", collapse = TRUE, eval = FALSE,
  dev = "png",
  dev.args = if (capabilities("cairo")) list(type = "cairo") else NULL
)
options(pkgdown.internet = FALSE)
```

## Introduction

**What is ksTFL?**

ksTFL is a lightweight R toolkit for building metadata specifications for clinical Tables, Figures, and Text (TFLs). Unlike traditional table formatters (flextable, huxtable, officer), ksTFL separates *metadata generation* from *rendering*:
- **You define**: Document structure, content, column formats, and styles in R
- **ksTFL generates**: A structured specification (JSON metadata + data files)
- **Built-in C++ renderer**: Produces submission-quality DOCX documents with deterministic HarfBuzz-based pagination via `write_doc()` / `replay_report()`

This separation enables:
- **Consistency**: All tables follow the same style templates
- **Automation**: Generate 50+ tables with consistent formatting and minimal code
- **Reproducibility**: Specifications are version-controllable and auditable
- **Flexibility**: Swap renderers without changing your R code

**When to use ksTFL:**
- Regulatory clinical reporting (FDA, EMA submissions)
- Large-scale automated reporting (100+ tables from common data sources)
- Complex styling requirements (multi-level headers, spanning columns, conditional formatting)
- When you need JSON metadata for downstream processing

## How to use this guide

This is the foundation vignette and best starting point for new users.

- Audience: users building first end-to-end TFL workflows
- Focus: core objects (`TFL_spec`, `TFL_report`), function flow, and minimal working examples
- Outcome: ability to create, assemble, and render reports confidently

Recommended reading order after this vignette:
1. `Reporting_Examples_with_ksTFL` for realistic workflow patterns.
2. `Styling_Guide_with_ksTFL` for style primitives and reusable style systems.
3. `Column_Width_Management` for layout tuning.
4. `Advanced_StyleRows` for conditional row actions.

---

## General workflow

### High-level pipeline

```
Data Frame / Image File
       ↓
create_table() / create_figure() / create_text()
       ↓
TFL_spec (customize with define_cols(), add_style(), add_title(), compute_cols())
       ↓
create_report() (assemble specs, consolidate styles)
       ↓
TFL_report
       ↓
write_doc()  ← recommended: one-step save + render
  or
save_report() → replay_report()  ← two-step: inspect JSON before rendering
       ↓
Styled .docx (C++ engine, deterministic HarfBuzz pagination)
```

### Anatomy of a TFL spec

Each `TFL_spec` contains:

| Component | Purpose | How to modify |
|-----------|---------|---------------|
| **document** | Metadata (docType, title, page settings) | `set_document()`, `set_page_style()` |
| **columns** | Column definitions, labels, formats, styles | `define_cols()` |
| **stubColumns** | Spanning headers above column groups | `add_span_header()` |
| **headers** | Page header text (left/center/right) | `add_header()` |
| **titles** | Main document titles | `add_title()` |
| **subtitles** | Secondary titles | `add_subtitle()` |
| **bodyText** | Narrative content | `add_body_text()` |
| **footnotes** | Document footnotes | `add_footnote()` |
| **footers** | Page footer text | `add_footer()` |
| **styles** | Named styles defined with `add_style()` | `add_style()` |

---

## Quick 5-minute example

Get a working spec in 5 minutes:

```{r quick_pipeline, eval = FALSE}
library(ksTFL)

# 1. Create a table spec from data
spec <- create_table(mtcars, cols = c(mpg, cyl, hp))

# 2. Define a style and apply it
spec <- add_style(spec, id = "header_bold",
  s_font(bold = TRUE, font_size = "12pt"))

# 3. Customize columns
spec <- define_cols(spec, c(mpg, cyl, hp),
  label = c("MPG", "Cylinders", "HP"),
  type = c("numeric", "numeric", "numeric"),
  format = c("%.1f", "%.0f", "%.0f"),
  labelStyleRef = "header_bold")

# 4. Add titles
spec <- add_title(spec, "Motor Trends Dataset")
spec <- add_footnote(spec, "Data from mtcars (1974).")

# 5. Inspect the spec
print(spec)

# 6. Render to DOCX
report <- create_report(spec)
write_doc(report, name = "my_table", outDir = "./output", metaPath = tempdir())
```

---

## Step 1: Create a TFL_spec

Choose the document type and initialize:

### Table spec (from data frame)

```{r create_table_spec, eval = FALSE}
# Simple: all columns
spec_tbl <- create_table(mtcars)

# Select specific columns with tidyselect
spec_tbl <- create_table(mtcars, cols = c(mpg, cyl, hp, wt))

```

**What happens**: 
- Data is copied into the spec's internal environment (not modifiable)
- Columns are auto-analyzed (type detection, width calculation)
- A `TFL_spec` object is returned ready for customization

### Figure spec (from image file)

```{r create_figure_spec, eval = FALSE}
# From a file path — any image format your renderer supports (PNG, JPEG, etc.)
spec_fig <- create_figure("path/to/plot.png")

# From a ggplot2 object — rendered automatically to a temporary PNG
library(ggplot2)
p <- ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()
spec_fig <- create_figure(p)                         # default 6x4in, 300 dpi
spec_fig <- create_figure(p, width = 8, height = 5) # custom dimensions
spec_fig <- create_figure(p, device = "jpeg")        # JPEG output
```

**What happens**:
- If a **file path** is given: it is validated (must exist and be readable); the file is used as-is
- If a **ggplot2 object** is given: rendered via `ggplot2::ggsave()` to `tempdir()` using the specified `device` (default `"png"`); the resulting path is stored in the spec
- The `width`, `height`, and `dpi` parameters only apply when a ggplot2 object is passed; they are ignored for file paths
- The image file is copied to `metaPath` when the report is saved

### Text spec (narrative content)

```{r create_text_spec, eval = FALSE}
spec_txt <- create_text()

# Add narrative content
spec_txt <- add_body_text(spec_txt, "This analysis includes all subjects in the safety population.")
```

**What happens**:
- Empty spec with no data
- Ready for narrative content via `add_body_text()`

---

## Step 2: Customize columns (define_cols)

After creating a table spec, customize columns with `define_cols()`:

### Single column

```{r define_single_column, eval = FALSE}
spec <- define_cols(spec, mpg,
  label = "Miles per Gallon",
  type = "numeric",
  format = "%.1f",
  colWidth = "20%")
```

### Batch update (multiple columns with single values)

```{r define_batch_single, eval = FALSE}
# Apply same label format to all columns
spec <- define_cols(spec, c(mpg, cyl, hp),
  label = c("MPG", "Cylinders", "HP"),
  type = "numeric",        # Recycled to all 3
  format = "%.0f")         # Recycled to all 3
```

### Per-column customization

```{r define_batch_mapped, eval = FALSE}
# Different format for each column
spec <- define_cols(spec, c(mpg, cyl, hp),
  label = c("MPG", "Cylinders", "HP"),
  type = c("numeric", "numeric", "numeric"),
  format = c("%.1f", "%.0f", "%.0f"))
```

**Key parameters**:
- `label`: Column header text
- `type`: "numeric" or "string" (auto-detected if omitted)
- `format`: Format string (e.g., "%.2f", "0.00")
- `colWidth`: Column width (e.g., "20%", "2cm")
  - When set, column is locked to that width
  - Other unlocked columns auto-recalculate to sum to 100%
  - Disable auto-recalculation with `tfl_set_options(autoColWidth = FALSE)`
- `isVisible`: TRUE (default) or FALSE
  - Set to FALSE to hide column from output
  - Invisible columns show width "0.0cm" 
  - Cannot set colWidth on invisible columns
  - Invisible columns excluded from width recalculation
- `isID`: TRUE if column repeats on page breaks
- `labelStyleRef`: Style(s) to apply to column header
- `valueStyleRef`: Style(s) to apply to cell values

**Hiding Columns**:
```r
# Hide a column (e.g., for conditional logic that doesn't display)
spec <- create_table(data) |>
  define_cols(group_id, isVisible = FALSE)
  # Result: group_id is hidden, other columns recalculated to fill 100%
```

See [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd) for detailed column customization patterns.

---

## Step 3: Define styles (add_style)

Create reusable named styles with `add_style()`:

```{r define_styles, eval = FALSE}
spec <- create_table(mtcars)

# Header style: bold, 12pt, centered, gray background
spec <- add_style(spec, id = "table_header",
  s_font(font_name = "Arial", font_size = "12pt", bold = TRUE),
  s_paragraph(alignment = "center"),
  s_table_style(background_color = "#E0E0E0"))

# Numeric style: right-aligned
spec <- add_style(spec, id = "numeric_right",
  s_paragraph(alignment = "right"))

# Apply to columns
spec <- define_cols(spec, c(mpg, hp),
  labelStyleRef = "table_header",
  valueStyleRef = "numeric_right")
```

**Best practice**: Define styles once, reference by id (name) throughout your spec.

For comprehensive styling details see [Styling Guide](Styling_Guide_with_ksTFL.Rmd).

---

## Step 4: Add spanning headers (span headers)

Create multi-level headers with `add_span_header()` using tidyselect expressions:

```{r add_stubs, eval = FALSE}
spec <- create_table(mtcars)
spec <- define_cols(spec, c(mpg, cyl, hp, wt),
  label = c("MPG", "Cyl", "HP", "Weight"))

# Top-level span header spanning all columns
spec <- add_span_header(spec, cols = c(mpg, cyl, hp, wt),
  label = "Motor Vehicle Specs",
  stubOrder = 0) |>
# Second-level headers: finer grouping
add_span_header(cols = c(mpg, cyl, hp),
  label = "Engine",
  stubOrder = 1) |>
add_span_header(cols = wt,
  label = "Weight",
  stubOrder = 1)
```

**Tidyselect support**: `add_span_header()` accepts all tidyselect expressions:

```{r add_stubs_tidyselect, eval = FALSE}
# Using column ranges and helpers
spec <- add_span_header(spec, cols = mpg:hp, label = "First", stubOrder = 0) |>
  add_span_header(cols = starts_with("c"), label = "C-columns", stubOrder = 1) |>
  add_span_header(cols = -mpg, label = "Non-MPG", stubOrder = 2)
```

**Key concepts**:
- `cols`: Column names (from `spec$columns`) to span. Accepts:
  - **Unquoted names**: `c(mpg, cyl, hp)` 
  - **Quoted names**: `c("mpg", "cyl", "hp")`
  - **Ranges**: `mpg:hp`
  - **Helpers**: `starts_with("c")`, `contains("w")`, `matches("^m")`
  - **Negation**: `-mpg` (all columns except mpg)
- `label`: Spanning header text
- `stubOrder`: Vertical stacking order (0 = top, 1 = next, etc.; auto-generated if NULL)
- `labelStyleRef`: Style(s) for the header label

**Rules**:
- Stubs at the **same** `stubOrder` cannot share columns (prevents ambiguous headers)
- Stubs at **different** `stubOrder` values can overlap freely (enables hierarchical structure)
- Multiple stubs at the same order are allowed as long as their column sets don't overlap
- Use `add_style()` to style stub labels

See [Styling Guide](Styling_Guide_with_ksTFL.Rmd) for styling stubs.

---

## Step 5: Add titles, footnotes, headers, footers

Add document content layers:

```{r add_content, eval = FALSE}
spec <- create_table(mtcars)

# Titles and subtitles
spec <- add_title(spec, "Demographics")
spec <- add_subtitle(spec, "Motor Vehicle Analysis")

# Footnotes (document-level notes)
spec <- add_footnote(spec, "Values are from 1974 Motor Trend magazine.")

# Page headers (left/center/right) — three separate string arguments
spec <- add_header(spec, "Study ABC", "CONFIDENTIAL", "Page {page}")

# Page footers (left/center/right) — three separate string arguments
spec <- add_footer(spec, "Company", "Locked DB", "2025")

# Body text (narrative)
spec <- add_body_text(spec, "This analysis includes all subjects in the safety population.")
```

**Notes**:
- `add_header()` and `add_footer()` take up to 3 separate string arguments: left, center, right
- Placeholders like `{page}` and `{numpages}` are filled in by the renderer
- Multiple `add_header()` calls append additional header rows; use the `level` parameter to replace a specific row
- Multiple `add_footnote()` calls stack in order
- `add_title()` and `add_subtitle()` accept an optional `toclevel` parameter (integer, e.g. `1` or `2`). When set, the title is included in the Table of Contents generated by `write_doc(toc = TRUE)`. See `?add_title` for details.

---

## Step 6: Combine specs into a report

Assemble multiple specs with `create_report()`:

```{r create_report, eval = FALSE}
# Create multiple specs
spec_table <- create_table(mtcars)
spec_table <- add_title(spec_table, "Table 1: Demographics")

spec_text <- create_text()
spec_text <- add_body_text(spec_text, "Analysis performed in R.")

spec_fig <- create_figure("plots/histogram.png")

# Combine into single report
report <- create_report(spec_table, spec_text, spec_fig)

# Inspect combined report
print(report)
```

**What `create_report()` does**:
1. Flattens inputs (supports nested reports)
2. Validates structure
3. Consolidates combined styles (if any used `f_combine()`)
4. Assigns sequential `docOrder` (1, 2, 3, ...)
5. Creates `dataRef` names for new specs
6. Returns a `TFL_report` object

**Result**: A single report combining all specs in input order.

---

## Step 7: Apply conditional row actions (compute_cols)

Beyond global column styling, apply **conditional actions** to specific rows that match a condition:

```{r compute_cols_basic, eval = FALSE}
spec <- create_table(mtcars)

# Define a style for first group occurrences
spec <- add_style(spec, id = "group_header", s_font(bold = TRUE, color = "#0000FF"))

# Apply style conditionally: highlight first occurrence of each cyl group
spec <- compute_cols(spec, firstOf(cyl), 
  c_style(c(mpg, hp), styleRef = "group_header"))

# Add a separator row above first group
spec <- compute_cols(spec, firstOf(cyl), 
  c_addrow(pos = "above"))
```

**Common use cases**:
- **Highlight group headers**: Style first/last occurrence of a group
- **Add separators**: Insert empty rows between groups
- **Merge columns**: Create group headers by merging columns
- **Conditional formatting**: Style rows meeting a threshold (e.g., `value > 100`)

**Key action functions** (used inside `compute_cols()`):

| Function | Purpose | Example |
|----------|---------|---------|
| `c_style(cols, styleRef)` | Apply style to columns in matching rows | `c_style(c(mpg, hp), styleRef = "bold")` |
| `c_merge(cols, styleRef = NULL)` | Merge adjacent columns | `c_merge(c(col1, col2), styleRef = "header")` |
| `c_addrow(pos, value_from = NULL, styleRef = NULL)` | Insert row above/below | `c_addrow(pos = "above")` for empty separator |
| `c_pageBreak()` | Insert a page break at the matching row (no args) | `c_pageBreak()` |

**Condition syntax**:

```{r compute_cols_conditions, eval = FALSE}
# Direct column comparison
compute_cols(spec, cyl > 6, c_style(mpg, styleRef = "emphasize"))

# String matching
compute_cols(spec, Parameter == "Pulse", c_style(value, styleRef = "numeric"))

# Helper functions (firstOf, lastOf, everyNth)
compute_cols(spec, firstOf(group), c_style(c(col1, col2), styleRef = "group_header"))

# Combine conditions
compute_cols(spec, cyl == 8 & hp > 100, c_style(mpg, styleRef = "high_power"))
```

**Key concepts**:
- Conditions are evaluated during `create_report()` — at that point, all data is available
- Multiple `compute_cols()` calls accumulate on the same spec
- Actions on the same row from different `compute_cols()` blocks are combined (styles merge automatically)
- `value_from = NULL` in `c_addrow()` creates an empty separator row

For detailed examples see [Reporting Examples: Conditional Row Actions](Reporting_Examples_with_ksTFL.Rmd#conditional-row-actions).

---

## Step 8: Save and Render

### Option A: One step with `write_doc()` (recommended)

`write_doc()` combines save + render into a single call:

```{r write_doc, eval = FALSE}
report <- create_report(spec_table, spec_text)

# Returns the full path to the generated .docx (invisibly)
doc_path <- write_doc(report,
  name     = "demographics",          # Output file name (no extension)
  outDir   = "./output",              # Where the final .docx is written
  metaPath = tempdir())               # Where JSON metadata is written (temp is fine)
```

**Additional parameters**:
- `toc`: If `TRUE`, prepends a Table of Contents page (requires `toclevel` on titles — see `?add_title`)
- `tocTitle`: Heading above the TOC field (default `"Table of Contents"`)
- `prettify`: If `TRUE`, writes human-readable JSON (useful for debugging)
- `verbose`: If `TRUE`, prints renderer progress messages

This is the recommended approach for production use. Set session defaults once:

```{r write_doc_with_options, eval = FALSE}
tfl_set_options(
  output_directory = "./output",
  meta_directory   = "./meta"
)

# Then write_doc() uses those defaults automatically
write_doc(report, name = "demographics")
```

### Option B: Two steps with `save_report()` + `replay_report()`

Use this when you need to inspect or version-control the JSON metadata separately:

```{r save_report, eval = FALSE}
report <- create_report(spec_table, spec_text)

# Step 1: Serialize to JSON + data files
result <- save_report(report,
  docFileName = "my_report.docx",
  outDir      = "./output",
  metaPath    = tempdir(),
  prettify    = TRUE)              # prettify = TRUE for readable JSON (debugging)

# Step 2: Re-render the DOCX from the saved JSON metadata
replay_report(
  spec_json   = result$spec_file,
  meta_dir    = result$metaPath,
  output_path = file.path("./output", "my_report.docx"))
```

**`save_report()` output files**:
- `{metaPath}/{spec_hash}.json` — Main specification (consumed by renderer)
- `{metaPath}/{dataRef}.json` — Table data files (one per table spec)
- Figure files copied to `metaPath` as-is

---

## Step 9: JSON metadata — what it is and why you might need it

### What is the metadata folder?

Every time `write_doc()` (or `save_report()`) runs, it writes intermediate files to `metaPath`:

- **`{hash}.json`** — The main spec file: document structure, column definitions, styles, titles, footnotes, headers/footers. This is what the C++ renderer reads.
- **`{dataRef}.json`** — One data file per table spec, containing the actual row data.
- **Figure files** — Copied from their original paths into `metaPath`.
- **`_index.json`** — An index of all spec files written to this folder (auto-maintained).

For most workflows, `metaPath = tempdir()` is fine — you only care about the final `.docx`. But there are real reasons to use a persistent `metaPath`.

### Why use a persistent metaPath?

**Reproducibility / audit trail**: In regulated environments (FDA, EMA submissions), you may need to prove that a specific DOCX was generated from a specific dataset at a specific time. The spec JSON + data JSON together form a complete, reproducible snapshot.

**Re-rendering without R**: Once the JSON files exist, you can re-render the DOCX without any R objects or data frames — just the JSON files. This is useful for:
- Regenerating a document after a template change
- Rendering on a different machine or in a CI/CD pipeline
- Reproducing an output months later

**Version management**: Each `write_doc()` call writes a new hash-named spec file. You can keep multiple versions and compare them.

### Managing metadata files

```{r metadata_management, eval = FALSE}
meta_dir <- "./meta"

# List all spec JSONs in the meta folder
df <- list_reports(meta_dir)
print(df[df$is_latest, c("doc_file", "datetime", "spec_file")])

# Re-render a DOCX from stored JSON (no R objects needed)
replay_report("demographics.docx", meta_dir = meta_dir)

# Or replay a specific version by spec hash
replay_report("abc123def456.json", meta_dir = meta_dir,
              output_path = "./output/demographics_v2.docx")

# Clean up obsolete spec files (keeps latest version per document)
clean_reports(meta_dir, dry_run = TRUE)   # Preview what would be deleted
clean_reports(meta_dir, dry_run = FALSE)  # Actually delete
```

**`list_reports()`** returns a data frame with columns: `doc_file`, `datetime`, `is_latest`, `n_specs`, `spec_file`, `data_refs`.

**`replay_report()`** re-renders from JSON — no R spec objects, no data frames required. Pass either the target `.docx` name (re-renders the latest version) or the exact spec hash filename.

**`clean_reports()`** removes obsolete JSON files. By default keeps 1 version per document (`keep_versions = 1`). Always run with `dry_run = TRUE` first.

### Recommended production setup

```{r production_setup, eval = FALSE}
# Set persistent meta directory once per session
tfl_set_options(
  output_directory = "./output",
  meta_directory   = "./meta"   # Persistent — survives session restarts
)

# All write_doc() calls now use these directories automatically
write_doc(report1, name = "tbl_demographics")
write_doc(report2, name = "tbl_labs")

# Later: re-render without R if needed
replay_report("tbl_demographics.docx", meta_dir = "./meta")
```

---

## Understanding auto-generated values

### Column widths (auto-calculation)

When you create a table, widths are auto-calculated based on data characteristics:

```r
# Auto-calculated widths (example)
# id = 20%, description = 50%, value = 30%
spec <- create_table(my_data)

# Lock id to 15%; others recalculate to fill 85%
spec <- define_cols(spec, id, colWidth = "15%")
```

See [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd) for detailed width customization.

### Data references (dataRef)

Each table spec gets a `dataRef` name for its data file:

```r
# Auto-generated: "0001_abc123def456" (docOrder + hash)
report <- create_report(spec_table1, spec_table2)

# Manual control not needed — handled automatically by write_doc()
write_doc(report, name = "my_report", outDir = "./output", metaPath = "./meta")
```

---

## Session-wide defaults with `tfl_options`

Set defaults once, inherited by all new specs:

```{r session_options, eval = FALSE}
# Set session defaults
tfl_set_options(
  add_header("Study ABC", "Locked Database", "CONFIDENTIAL"),
  add_footer("Company Name", "Page {page}", "2025"),
  add_body_text("Analysis performed in R with ksTFL."))

# All new specs created afterward inherit these settings
spec1 <- create_table(data1)  # Automatically gets headers/footers/body text
spec2 <- create_table(data2)  # Also inherits defaults

# Check current options
tfl_get_options()

# Reset to package defaults
tfl_reset_options()
```

---

## Complete example: From data to document

```{r complete_example, eval = FALSE}
library(ksTFL)

# --- Setup ---
data <- data.frame(
  subject = sprintf("S%03d", 1:10),
  age = round(rnorm(10, 45, 10)),
  sex = sample(c("M", "F"), 10, TRUE)
)

# --- Create & customize ---
spec <- create_table(data, cols = c(subject, age, sex))

spec <- add_style(spec, id = "header",
  s_font(bold = TRUE, font_size = "11pt"),
  s_paragraph(alignment = "center"),
  s_table_style(background_color = "#DDDDDD"))

spec <- define_cols(spec, c(subject, age, sex),
  label = c("Subject ID", "Age (years)", "Sex"),
  labelStyleRef = "header")

spec <- define_cols(spec, age, type = "numeric", format = "%.0f")

spec <- add_title(spec, "Demographics Table")
spec <- add_subtitle(spec, "All subjects in safety population")
spec <- add_footnote(spec, "Data are shown as observed.")

# --- Export ---
report <- create_report(spec)
write_doc(report, name = "demographics", outDir = "./output", metaPath = tempdir())

# Done!
```

---

## Key gotchas and tips

| Gotcha | Solution |
|--------|----------|
| **Calling s_* helpers outside add_style()** | Always use s_* inside `add_style()` — they validate context |
| **define_cols() with mismatched parameter lengths** | Length must be 1 (recycled) or match number of columns |
| **Styles from f_combine() not consolidated** | Call `create_report()` before `write_doc()` |
| **Multiple add_header() calls stack** | Each call **appends** a new header level; use `level =` to replace a specific level |
| **Figure file not found** | Check path is absolute or relative to current working directory |
| **create_text() with data argument** | `create_text()` requires `data = NULL` — no data for text specs |
| **Overlapping stubs at same stubOrder** | Error; use different `stubOrder` or non-overlapping column sets |

---

## Next steps

You now understand ksTFL's core workflow. Next steps:

1. **Run the quick example** above with your own data
2. **Explore [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd)** for detailed patterns
3. **Read [Styling Guide](Styling_Guide_with_ksTFL.Rmd)** for advanced styling
4. **Read [Advanced StyleRows](Advanced_StyleRows.Rmd)** for conditional formatting
5. **Check function docs**: `?create_table`, `?define_cols`, `?add_style`, `?write_doc`, `?list_reports`

---

## Resources

- **Function reference**: `?ksTFL` (package overview) or `?create_table`, `?define_cols`, etc.
- **Reporting Examples**: [Detailed working examples with explanations](Reporting_Examples_with_ksTFL.Rmd)
- **Styling Guide**: [Comprehensive style reference](Styling_Guide_with_ksTFL.Rmd)
- **GitHub**: [ksTFL repository](https://github.com/...)

---

## Quick reference table

| Task | Function | Notes |
|------|----------|-------|
| Create table spec | `create_table(data, cols = ...)` | Auto-detects columns, types, widths |
| Create figure spec | `create_figure(plot_or_path, ...)` | File path or ggplot2 object; path must be readable |
| Create text spec | `create_text()` | For narrative content only |
| Set document properties | `set_document(spec, hasData = ...)` | Optional: configure content width, placement |
| Customize columns | `define_cols(spec, cols, ...)` | Use `c()` for multiple columns |
| Define styles | `add_style(spec, id = "name", ...)` | Use `s_font()`, `s_paragraph()`, `s_table_style()` |
| Add spanning header | `add_span_header(spec, cols, label, ...)` | Supports tidyselect; multi-level headers |
| Add titles/content | `add_title()`, `add_footnote()`, `add_body_text()` | Layer document content |
| Add page headers/footers | `add_header()`, `add_footer()` | 3 parts: left/center/right |
| Set page style | `set_page_style(spec, page = p_page(...))` | Configure page size, orientation, margins |
| Combine specs | `create_report(spec1, spec2, ...)` | Consolidates styles, assigns order |
| Save + render (one step) | `write_doc(report, name, ...)` | Recommended: saves JSON + renders DOCX |
| Save only | `save_report(report, ...)` | Writes JSON + data files for manual rendering |
| Render from JSON | `replay_report(spec_json, meta_dir, output_path)` | C++ renderer: JSON → styled DOCX |
| Set session defaults | `tfl_set_options(...)` | Inherited by new specs in session |
| Check options | `tfl_get_options()`, `tfl_get_option(name)` | View current session settings |
| Reset options | `tfl_reset_options()` | Back to package defaults |
