---
title: "Getting Started with ksTFL"
author: "ksTFL Team"
output:
  rmarkdown::html_vignette:
    dev: ragg_png
vignette: >
  %\VignetteIndexEntry{Getting Started with ksTFL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", collapse = TRUE, eval = FALSE, dev = "ragg_png")
```

## Introduction

**What is ksTFL?**

ksTFL is a lightweight R toolkit for building metadata specifications for clinical Tables, Figures, and Text (TFLs). Unlike traditional table formatters (flextable, huxtable, officer), ksTFL separates *metadata generation* from *rendering*:
- **You define**: Document structure, content, column formats, and styles in R
- **ksTFL generates**: A structured specification (JSON metadata + data files)
- **Built-in C++ renderer**: Produces submission-quality DOCX documents with deterministic HarfBuzz-based pagination via `render_docx()`

This separation enables:
- **Consistency**: All tables follow the same style templates
- **Automation**: Generate 50+ tables with consistent formatting and minimal code
- **Reproducibility**: Specifications are version-controllable and auditable
- **Flexibility**: Swap renderers without changing your R code

**When to use ksTFL:**
- Regulatory clinical reporting (FDA, EMA submissions)
- Large-scale automated reporting (100+ tables from common data sources)
- Complex styling requirements (multi-level headers, spanning columns, conditional formatting)
- When you need JSON metadata for downstream processing

---

## General workflow

### High-level pipeline

```
Data Frame / Image File
       ↓
create_table() / create_figure() / create_text()
       ↓
TFL_spec (customize with define_cols(), add_style(), add_title())
       ↓
create_report() (assemble specs, consolidate styles)
       ↓
TFL_report
       ↓
save_report() (validate, serialize, write JSON + data files)
       ↓
render_docx() (C++ engine → deterministic pagination → styled .docx)
```

### Anatomy of a TFL spec

Each `TFL_spec` contains:

| Component | Purpose | How to modify |
|-----------|---------|---------------|
| **document** | Metadata (docType, title, page settings) | `set_document()`, `set_page_style()` |
| **columns** | Column definitions, labels, formats, styles | `define_cols()` |
| **stubColumns** | Spanning headers above column groups | `add_span_header()` |
| **headers** | Page header text (left/center/right) | `add_header()` |
| **titles** | Main document titles | `add_title()` |
| **subtitles** | Secondary titles | `add_subtitle()` |
| **bodyText** | Narrative content | `add_body_text()` |
| **footnotes** | Document footnotes | `add_footnote()` |
| **footers** | Page footer text | `add_footer()` |
| **styles** | Named styles defined with `add_style()` | `add_style()` |

---

## Quick 5-minute example

Get a working spec in 5 minutes:

```{r quick_pipeline, eval = FALSE}
library(ksTFL)

# 1. Create a table spec from data
spec <- create_table(mtcars, cols = c(mpg, cyl, hp))

# 2. Define a style and apply it
spec <- add_style(spec, id = "header_bold",
  s_font(bold = TRUE, font_size = "12pt"))

# 3. Customize columns
spec <- define_cols(spec, c(mpg, cyl, hp),
  label = c("MPG", "Cylinders", "HP"),
  type = c("numeric", "numeric", "numeric"),
  format = c("%.1f", "%.0f", "%.0f"),
  labelStyleRef = "header_bold")

# 4. Add titles
spec <- add_title(spec, "Motor Trends Dataset")
spec <- add_footnote(spec, "Data from mtcars (1974).")

# 5. Inspect the spec
print(spec)

# 6. Save for rendering
report <- create_report(spec)
save_report(report, docFileName = "my_table.docx", outDir = "./output", metaPath = tempdir())
```

---

## Step 1: Create a TFL_spec

Choose the document type and initialize:

### Table spec (from data frame)

```{r create_table_spec, eval = FALSE}
# Simple: all columns
spec_tbl <- create_table(mtcars)

# Select specific columns with tidyselect
spec_tbl <- create_table(mtcars, cols = c(mpg, cyl, hp, wt))

```

**What happens**: 
- Data is copied into the spec's internal environment (not modifiable)
- Columns are auto-analyzed (type detection, width calculation)
- A `TFL_spec` object is returned ready for customization

### Figure spec (from image file)

```{r create_figure_spec, eval = FALSE}
# From a file path (PNG, JPEG, SVG — any tool: R base, Python, Inkscape, etc.)
spec_fig <- create_figure("path/to/plot.png")

# From a ggplot2 object — rendered automatically to a temporary PNG
library(ggplot2)
p <- ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()
spec_fig <- create_figure(p)                         # default 6x4in, 300 dpi
spec_fig <- create_figure(p, width = 8, height = 5) # custom dimensions
spec_fig <- create_figure(p, device = "jpeg")        # JPEG output
```

**What happens**:
- If a **file path** is given: it is validated (must exist and be readable)
- If a **ggplot2 object** is given: rendered via `ggplot2::ggsave()` to `tempdir()`; the resulting path is stored in the spec
- Minimal metadata is created; the file is copied to the output directory when saved

### Text spec (narrative content)

```{r create_text_spec, eval = FALSE}
spec_txt <- create_text()

# Add narrative content
spec_txt <- add_body_text(spec_txt, "This analysis includes all subjects in the safety population.")
```

**What happens**:
- Empty spec with no data
- Ready for narrative content via `add_body_text()`

---

## Step 2: Customize columns (define_cols)

After creating a table spec, customize columns with `define_cols()`:

### Single column

```{r define_single_column, eval = FALSE}
spec <- define_cols(spec, mpg,
  label = "Miles per Gallon",
  type = "numeric",
  format = "%.1f",
  colWidth = "20%")
```

### Batch update (multiple columns with single values)

```{r define_batch_single, eval = FALSE}
# Apply same label format to all columns
spec <- define_cols(spec, c(mpg, cyl, hp),
  label = c("MPG", "Cylinders", "HP"),
  type = "numeric",        # Recycled to all 3
  format = "%.0f")         # Recycled to all 3
```

### Per-column customization

```{r define_batch_mapped, eval = FALSE}
# Different format for each column
spec <- define_cols(spec, c(mpg, cyl, hp),
  label = c("MPG", "Cylinders", "HP"),
  type = c("numeric", "numeric", "numeric"),
  format = c("%.1f", "%.0f", "%.0f"))
```

**Key parameters**:
- `label`: Column header text
- `type`: "numeric" or "string" (auto-detected if omitted)
- `format`: Format string (e.g., "%.2f", "0.00")
- `colWidth`: Column width (e.g., "20%", "2cm")
  - When set, column is locked to that width
  - Other unlocked columns auto-recalculate to sum to 100%
  - Disable with `tfl_set_options(autoColWidth = FALSE)`
- `isVisible`: TRUE (default) or FALSE
  - Set to FALSE to hide column from output
  - Invisible columns show width "0.0cm" 
  - Cannot set colWidth on invisible columns
  - Invisible columns excluded from width recalculation
- `isID`: TRUE if column repeats on page breaks
- `labelStyleRef`: Style(s) to apply to column header
- `valueStyleRef`: Style(s) to apply to cell values

**Hiding Columns**:
```r
# Hide a column (e.g., for conditional logic that doesn't display)
spec <- create_table(data) |>
  define_cols(group_id, isVisible = FALSE)
  # Result: group_id is hidden, other columns recalculated to fill 100%
```

See [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd) for detailed column customization patterns.

---

## Step 4: Define styles (add_style)

Create reusable named styles with `add_style()`:

```{r define_styles, eval = FALSE}
spec <- create_table(mtcars)

# Header style: bold, 12pt, centered, gray background
spec <- add_style(spec, id = "table_header",
  s_font(font_name = "Arial", font_size = "12pt", bold = TRUE),
  s_paragraph(alignment = "center"),
  s_table_style(background_color = "#E0E0E0"))

# Numeric style: right-aligned
spec <- add_style(spec, id = "numeric_right",
  s_paragraph(alignment = "right"))

# Apply to columns
spec <- define_cols(spec, c(mpg, hp),
  labelStyleRef = "table_header",
  valueStyleRef = "numeric_right")
```

**Best practice**: Define styles once, reference by id (name) throughout your spec.

For comprehensive styling details see [Styling Guide](Styling_Guide_with_ksTFL.Rmd).

---

## Step 5: Add spanning headers (span headers)

Create multi-level headers with `add_span_header()` using tidyselect expressions:

```{r add_stubs, eval = FALSE}
spec <- create_table(mtcars)
spec <- define_cols(spec, c(mpg, cyl, hp, wt),
  label = c("MPG", "Cyl", "HP", "Weight"))

# Top-level span header spanning all columns
spec <- add_span_header(spec, cols = c(mpg, cyl, hp, wt),
  label = "Motor Vehicle Specs",
  stubOrder = 0) |>
# Second-level headers: finer grouping
add_span_header(cols = c(mpg, cyl, hp),
  label = "Engine",
  stubOrder = 1) |>
add_span_header(cols = wt,
  label = "Weight",
  stubOrder = 1)
```

**Tidyselect support**: `add_span_header()` accepts all tidyselect expressions:

```{r add_stubs_tidyselect, eval = FALSE}
# Using column ranges and helpers
spec <- add_span_header(spec, cols = mpg:hp, label = "First", stubOrder = 0) |>
  add_span_header(cols = starts_with("c"), label = "C-columns", stubOrder = 1) |>
  add_span_header(cols = -mpg, label = "Non-MPG", stubOrder = 2)
```

**Key concepts**:
- `cols`: Column names (from `spec$columns`) to span. Accepts:
  - **Unquoted names**: `c(mpg, cyl, hp)` 
  - **Quoted names**: `c("mpg", "cyl", "hp")`
  - **Ranges**: `mpg:hp`
  - **Helpers**: `starts_with("c")`, `contains("w")`, `matches("^m")`
  - **Negation**: `-mpg` (all columns except mpg)
- `label`: Spanning header text
- `stubOrder`: Vertical stacking order (0 = top, 1 = next, etc.; auto-generated if NULL)
- `labelStyleRef`: Style(s) for the header label

**Rules**:
- Stubs at the **same** `stubOrder` cannot share columns (prevents ambiguous headers)
- Stubs at **different** `stubOrder` values can overlap freely (enables hierarchical structure)
- Multiple stubs at the same order are allowed as long as their column sets don't overlap
- Use `add_style()` to style stub labels

See [Styling Guide](Styling_Guide_with_ksTFL.Rmd) for styling stubs.

---

## Step 6: Add titles, footnotes, headers, footers

Add document content layers:

```{r add_content, eval = FALSE}
spec <- create_table(mtcars)

# Titles and subtitles
spec <- add_title(spec, "Demographics")
spec <- add_subtitle(spec, "Motor Vehicle Analysis")

# Footnotes (document-level notes)
spec <- add_footnote(spec, "Values are from 1974 Motor Trend magazine.")

# Page headers (left/center/right)
spec <- add_header(c("Study ABC", "CONFIDENTIAL", "Page {page}"))

# Page footers (left/center/right)
spec <- add_footer(c("Company", "Locked DB", "2025"))

# Body text (narrative)
spec <- add_body_text(spec, "This analysis includes all subjects in the safety population.")
```

**Notes**:
- `add_header()` and `add_footer()` take 3 arguments: left-aligned, center-aligned, right-aligned
- Placeholders like `{page}` are filled in by the renderer
- Each call **replaces** previous headers/footers (does not append)
- Multiple `add_footnote()` calls stack

---

## Step 7: Combine specs into a report

Assemble multiple specs with `create_report()`:

```{r create_report, eval = FALSE}
# Create multiple specs
spec_table <- create_table(mtcars)
spec_table <- add_title(spec_table, "Table 1: Demographics")

spec_text <- create_text()
spec_text <- add_body_text(spec_text, "Analysis performed in R.")

spec_fig <- create_figure("plots/histogram.png")

# Combine into single report
report <- create_report(spec_table, spec_text, spec_fig)

# Inspect combined report
print(report)
```

**What `create_report()` does**:
1. Flattens inputs (supports nested reports)
2. Validates structure
3. Consolidates combined styles (if any used `f_combine()`)
4. Assigns sequential `docOrder` (1, 2, 3, ...)
5. Creates `dataRef` names for new specs
6. Returns a `TFL_report` object

**Result**: A single report combining all specs in input order.

---

## Step 8: Apply conditional row actions (compute_cols)

Beyond global column styling, apply **conditional actions** to specific rows that match a condition:

```{r compute_cols_basic, eval = FALSE}
spec <- create_table(mtcars)

# Define a style for first group occurrences
spec <- add_style(spec, id = "group_header", s_font(bold = TRUE, color = "#0000FF"))

# Apply style conditionally: highlight first occurrence of each cyl group
spec <- compute_cols(spec, firstOf(cyl), 
  c_style(c(mpg, hp), styleRef = "group_header"))

# Add a separator row above first group
spec <- compute_cols(spec, firstOf(cyl), 
  c_addrow(pos = "above"))
```

**Common use cases**:
- **Highlight group headers**: Style first/last occurrence of a group
- **Add separators**: Insert empty rows between groups
- **Merge columns**: Create group headers by merging columns
- **Conditional formatting**: Style rows meeting a threshold (e.g., `value > 100`)

**Key action functions** (used inside `compute_cols()`):

| Function | Purpose | Example |
|----------|---------|---------|
| `c_style(cols, styleRef)` | Apply style to columns in matching rows | `c_style(c(mpg, hp), styleRef = "bold")` |
| `c_merge(cols, styleRef = NULL)` | Merge adjacent columns | `c_merge(c(col1, col2), styleRef = "header")` |
| `c_addrow(pos, value_from = NULL, styleRef = NULL)` | Insert row above/below | `c_addrow(pos = "above")` for empty separator |
| `c_pageBreak()` | Insert a page break at the matching row (no args) | `c_pageBreak()` |

**Condition syntax**:

```{r compute_cols_conditions, eval = FALSE}
# Direct column comparison
compute_cols(spec, cyl > 6, c_style(mpg, styleRef = "emphasize"))

# String matching
compute_cols(spec, Parameter == "Pulse", c_style(value, styleRef = "numeric"))

# Helper functions (firstOf, lastOf, everyNth)
compute_cols(spec, firstOf(group), c_style(c(col1, col2), styleRef = "group_header"))

# Combine conditions
compute_cols(spec, cyl == 8 & hp > 100, c_style(mpg, styleRef = "high_power"))
```

**Key concepts**:
- Conditions are evaluated during `create_report()` — at that point, all data is available
- Multiple `compute_cols()` calls accumulate on the same spec
- Actions on the same row from different `compute_cols()` blocks are combined (styles merge automatically)
- `value_from = NULL` in `c_addrow()` creates an empty separator row

For detailed examples see [Reporting Examples: Conditional Row Actions](Reporting_Examples_with_ksTFL.Rmd#conditional-row-actions).

---

## Step 9: Save for Rendering

Export to JSON metadata + data files:

```{r save_report, eval = FALSE}
report <- create_report(spec_table, spec_text)

# Save with validation and serialization
result <- save_report(report,
  docFileName = "my_report.docx",
  outDir = "./output",
  metaPath = tempdir(),
  prettify = TRUE)

# Inspect output
str(result)
```

**Parameters**:
- `report`: A `TFL_report` object
- `docFileName`: Output file name (e.g., "report.docx")
- `outDir`: Directory for final DOCX (where renderer writes output)
- `metaPath`: Directory for JSON metadata + data files (consumed by renderer)
- `prettify`: If TRUE, formats JSON nicely (useful for debugging)

**Output files**:
- `{metaPath}/{spec_hash}.json` — Main specification (consumed by renderer)
- `{metaPath}/{dataRef}.json` — Table data files (one per table spec)
- Figure files copied to `metaPath` as-is

**Return value**: List with:
- `spec_file`: Path to generated spec JSON
- `data_files`: Named list of data file paths
- `metadata`: Report metadata (timestamps, etc.)

---

## Understanding auto-generated values

### Column widths (auto-calculation)

When you create a table, widths are auto-calculated based on data characteristics:

```r
# Auto-calculated widths (example)
# id = 20%, description = 50%, value = 30%
spec <- create_table(my_data)

# Lock id to 15%; others recalculate to fill 85%
spec <- define_cols(spec, id, colWidth = "15%")
```

See [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd) for detailed width customization.

### Data references (dataRef)

Each table spec gets a `dataRef` name for its data file:

```r
# Auto-generated: "0001_abc123def456" (docOrder + hash)
report <- create_report(spec_table1, spec_table2)

# Manual control not needed — handled automatically
save_report(report, ...)
```

---

## Session-wide defaults with `tfl_options`

Set defaults once, inherited by all new specs:

```{r session_options, eval = FALSE}
# Set session defaults
tfl_set_options(
  add_header(c("Study ABC", "Locked Database", "CONFIDENTIAL")),
  add_footer(c("Company Name", "Page {page}", "2025")),
  add_body_text("Analysis performed in R with ksTFL."))

# All new specs created afterward inherit these settings
spec1 <- create_table(data1)  # Automatically gets headers/footers/body text
spec2 <- create_table(data2)  # Also inherits defaults

# Check current options
tfl_get_options()

# Reset to package defaults
tfl_reset_options()
```

---

## Complete example: From data to document

```{r complete_example, eval = FALSE}
library(ksTFL)

# --- Setup ---
data <- data.frame(
  subject = sprintf("S%03d", 1:10),
  age = round(rnorm(10, 45, 10)),
  sex = sample(c("M", "F"), 10, TRUE)
)

# --- Create & customize ---
spec <- create_table(data, cols = c(subject, age, sex))

spec <- add_style(spec, id = "header",
  s_font(bold = TRUE, font_size = "11pt"),
  s_paragraph(alignment = "center"),
  s_table_style(background_color = "#DDDDDD"))

spec <- define_cols(spec, c(subject, age, sex),
  label = c("Subject ID", "Age (years)", "Sex"),
  labelStyleRef = "header")

spec <- define_cols(spec, age, type = "numeric", format = "%.0f")

spec <- add_title(spec, "Demographics Table")
spec <- add_subtitle(spec, "All subjects in safety population")
spec <- add_footnote(spec, "Data are shown as observed.")

# --- Export ---
report <- create_report(spec)
save_report(report, 
  docFileName = "demographics.docx",
  outDir = "./output",
  metaPath = tempdir())

# Done!
```

---

## Key gotchas and tips

| Gotcha | Solution |
|--------|----------|
| **Calling s_* helpers outside add_style()** | Always use s_* inside `add_style()` — they validate context |
| **define_cols() with mismatched parameter lengths** | Length must be 1 (recycled) or match number of columns |
| **Styles from f_combine() not consolidated** | Call `create_report()` before `save_report()` |
| **Multiple add_header() calls stack** | Each call **replaces** previous headers (last-win) |
| **Figure file not found** | Check path is absolute or relative to current working directory |
| **create_text() with data argument** | `create_text()` requires `data = NULL` — no data for text specs |
| **Overlapping stubs at same stubOrder** | Error; use different `stubOrder` or non-overlapping column sets |

---

## Next steps

You now understand ksTFL's core workflow. Next steps:

1. **Run the quick example** above with your own data
2. **Explore [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd)** for detailed patterns
3. **Read [Styling Guide](Styling_Guide_with_ksTFL.Rmd)** for advanced styling
4. **Check function docs**: `?create_table`, `?define_cols`, `?add_style`, `?create_report`

---

## Resources

- **Function reference**: `?ksTFL` (package overview) or `?create_table`, `?define_cols`, etc.
- **Reporting Examples**: [Detailed working examples with explanations](Reporting_Examples_with_ksTFL.Rmd)
- **Styling Guide**: [Comprehensive style reference](Styling_Guide_with_ksTFL.Rmd)
- **GitHub**: [ksTFL repository](https://github.com/...)

---

## Quick reference table

| Task | Function | Notes |
|------|----------|-------|
| Create table spec | `create_table(data, cols = ...)` | Auto-detects columns, types, widths |
| Create figure spec | `create_figure(plot_or_path, ...)` | File path or ggplot2 object; path must be readable |
| Create text spec | `create_text()` | For narrative content only |
| Set document properties | `set_document(spec, hasData = ...)` | Optional: configure content width, placement |
| Customize columns | `define_cols(spec, cols, ...)` | Use `c()` for multiple columns |
| Define styles | `add_style(spec, id = "name", ...)` | Use `s_font()`, `s_paragraph()`, `s_table_style()` |
| Add spanning header | `add_span_header(spec, cols, label, ...)` | Supports tidyselect; multi-level headers |
| Add titles/content | `add_title()`, `add_footnote()`, `add_body_text()` | Layer document content |
| Add page headers/footers | `add_header()`, `add_footer()` | 3 parts: left/center/right |
| Set page style | `set_page_style(spec, page = p_page(...))` | Configure page size, orientation, margins |
| Combine specs | `create_report(spec1, spec2, ...)` | Consolidates styles, assigns order |
| Export | `save_report(report, ...)` | Writes JSON + data files |
| Set session defaults | `tfl_set_options(...)` | Inherited by new specs in session |
| Check options | `tfl_get_options()`, `tfl_get_option(name)` | View current session settings |
| Reset options | `tfl_reset_options()` | Back to package defaults |
