---
title: "Reporting Examples — ksTFL (v3)"
author: "ksTFL Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reporting Examples — ksTFL (v3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette presents a realistic progression of reporting examples — from minimal tables/listings/figures/text to a full clinical-style table with multi-level headers, spanning stub columns, multi-line titles/footnotes, reusable styles, and session defaults. All examples use exported ksTFL helpers only; set `eval=TRUE` locally to run the R chunks.

```{r create_sample_data, eval = FALSE}
# Sample datasets used by multiple sections (run locally before executing examples)
set.seed(2025)

demog_tbl <- data.frame(
  subject_id = sprintf("S%03d", 1:24),
  age = sample(25:80, 24, TRUE),
  sex = sample(c("M","F"), 24, TRUE),
  trt = sample(c("Placebo","DrugA"), 24, TRUE),
  stringsAsFactors = FALSE
)

vitals_tbl <- data.frame(
  subject_id = rep(demog_tbl$subject_id, each = 2),
  visit = rep(c("Baseline","Week 12"), times = 24),
  sbp = round(rnorm(48, 120, 12)),
  dbp = round(rnorm(48, 75, 8)),
  stringsAsFactors = FALSE
)

labs_tbl <- data.frame(
  subject_id = demog_tbl$subject_id,
  ALT = round(rnorm(24, 28, 9)),
  AST = round(rnorm(24, 26, 8)),
  stringsAsFactors = FALSE
)

# Example plot file (created locally when running the vignette)
plot_file <- file.path(tempdir(), "example_plot.png")
png(plot_file, width = 600, height = 400)
plot(demog_tbl$age, vitals_tbl$sbp[1:24], xlab = "Age", ylab = "SBP", main = "Age vs SBP")
dev.off()
```

## 1 — Simple minimal table

Goal: show a single-table workflow from data to a `TFL_spec` using only exported helpers.

```{r example_minimal_table, eval = FALSE}
spec_min_table <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt), id = "tbl_min")
spec_min_table <- add_title(spec_min_table, "Demographics (minimal)")

# Print shows a compact overview (useful in interactive sessions)
print(spec_min_table)

```{r save_min_table, eval = FALSE}
# End-to-end: wrap single spec into a report and save
rpt_min_table <- create_report(spec_min_table)
res_min_table <- save_report(rpt_min_table, docFileName = "tbl_min.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_min_table
```
```

Notes: `print()` on a `TFL_spec` provides a readable overview including docType, titles and defined columns.

---

## 2 — Simple minimal figure

Goal: create a figure spec from a local image file and add a title.

```{r example_minimal_figure, eval = FALSE}
spec_min_fig <- create_figure(data = plot_file, id = "fig_min")
spec_min_fig <- add_title(spec_min_fig, "Example: Age vs SBP")
print(spec_min_fig)

```{r save_min_fig, eval = FALSE}
# End-to-end: save the single-figure report
rpt_min_fig <- create_report(spec_min_fig)
res_min_fig <- save_report(rpt_min_fig, docFileName = "fig_min.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_min_fig
```
```

---

## 3 — Simple minimal text (narrative)

Goal: create a text-only spec for narrative content.

```{r example_minimal_text, eval = FALSE}
spec_min_text <- create_text(id = "txt_min")
spec_min_text <- add_body_text(spec_min_text, "This narrative describes the study population and analysis approach.")
print(spec_min_text)

```{r save_min_text, eval = FALSE}
# End-to-end: save the narrative as a report
rpt_min_text <- create_report(spec_min_text)
res_min_text <- save_report(rpt_min_text, docFileName = "txt_min.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_min_text
```
```

---

## 4 — Combine table/figure/text into a single report

Goal: assemble mixed specs into a `TFL_report` and inspect its summary.

```{r example_combine_report, eval = FALSE}
report_simple <- create_report(spec_min_table, spec_min_fig, spec_min_text)
print(report_simple)

```{r save_report_simple, eval = FALSE}
# Save combined simple report
res_report_simple <- save_report(report_simple, docFileName = "report_simple.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_report_simple
```
```

Notes: `create_report()` preserves input order and consolidates styles and ids.

---

## 5 — Table with multiple levels (headers/subheaders), titles and footnotes

Goal: build a clinical-style table with multi-level column headers, subtitles and footnotes.

```{r example_multilevel_table, eval = FALSE}
spec_multi <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST), id = "tbl_multi")
spec_multi <- add_title(spec_multi, "Laboratory Results")
spec_multi <- add_subtitle(spec_multi, "Selected hepatic enzymes by subject")
spec_multi <- add_footnote(spec_multi, "Values are shown as observed.")

spec_multi <- define_cols(
  spec_multi,
  c(subject_id, ALT, AST),
  label = c("Subject ID", "ALT (U/L)", "AST (U/L)")
)

print(spec_multi)

```{r save_multilevel, eval = FALSE}
# Save the multilevel table as a report
rpt_multi <- create_report(spec_multi)
res_multi <- save_report(rpt_multi, docFileName = "tbl_multi.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_multi
```
```

---

## 6 — Stub columns (spanning headers)

Goal: demonstrate `add_stub_column()` to create stacked/spanning stub headers above groups of columns.

```{r example_stubs, eval = FALSE}
# Start with a simple table that contains multiple measurement columns
spec_stub <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt), id = "tbl_stub")
spec_stub <- define_cols(spec_stub, c(subject_id, age, sex, trt), label = c("Subject","Age","Sex","Treatment"))

# Add a first-level stub that groups age and sex as 'Vitals'
spec_stub <- add_stub_column(spec_stub, cols = c("age", "sex"), label = "Vitals", stubOrder = 1)

# Add a second-level stub that groups treatment separately
spec_stub <- add_stub_column(spec_stub, cols = c("trt"), label = "Treatment group", stubOrder = 2, labelStyleRef = "stub_label")

print(spec_stub)
```

Notes: `add_stub_column()` takes `cols` as column *names* from `spec$columns`, a `label`, and a numeric `stubOrder` (lower numbers render closer to the left/top depending on renderer). Provide `labelStyleRef` to style stub labels.

---

```{r save_stub, eval = FALSE}
# Save stubbed table report
rpt_stub <- create_report(spec_stub)
res_stub <- save_report(rpt_stub, docFileName = "tbl_stub.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_stub
```

## 7 — Styles and `f_combine()` for combined style references

Goal: create simple styles, combine them with `f_combine()` and reference the combined style in `define_cols()`.

```{r example_styles_combine, eval = FALSE}
# Create a spec and add some atomic styles
spec_styles <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt), id = "tbl_styles")
spec_styles <- add_style(spec_styles, id = "header_bold", s_font(bold = TRUE))
spec_styles <- add_style(spec_styles, id = "red_text", s_font(color = "#FF0000"))

# Combine two style refs into one labelStyleRef using f_combine()
spec_styles <- define_cols(
  spec_styles,
  c(subject_id, age, sex, trt),
  label = c("Subject","Age","Sex","Treatment"),
  labelStyleRef = c(f_combine("header_bold","red_text"), "", "", "")
)

print(spec_styles)
```

Notes: `f_combine()` returns a style reference that instructs the package to merge named styles into a single applied style on render/assembly.

---

```{r save_styles, eval = FALSE}
# Save the styled table report
rpt_styles <- create_report(spec_styles)
res_styles <- save_report(rpt_styles, docFileName = "tbl_styles.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_styles
```

## 8 — Save report (end-to-end export)

Goal: show `save_report()` usage — it validates and serializes specs and writes metadata/data/figure files to `metaPath`.

```{r example_save_report, eval = FALSE}
# Combine a couple of specs into a report
report_full <- create_report(spec_styles, spec_multi)

# Save the report — this calls the serializer internally and writes files to metaPath
res <- save_report(report_full, docFileName = "example_report.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)

# Inspect returned metadata
str(res)
```

Notes: `save_report()` is the public entrypoint to persist specs/data/figures for downstream rendering; it internally calls validation/serialization and returns a metadata list with generated file paths.

---

## 9 — Session-wide `tfl_options`: common headers/footers and default body text

Goal: set session defaults so multiple specs inherit common headers, footers, and default body text.

```{r example_session_options, eval = FALSE}
session_opts <- tfl_set_options(
  add_header(c("Study ABC", "Locked Database")),
  add_footer(c("Confidential", "Page {page}")),
  add_body_text("This report contains preliminary analyses.")
)

spec_with_opts <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt), id = "tbl_opts")
spec_with_opts <- add_title(spec_with_opts, "Demographics (session defaults applied)")
print(spec_with_opts)
```

Notes: Use `tfl_get_options()` to inspect current session defaults. Options apply only to specs created after calling `tfl_set_options()` in the same R session.

---

## Final notes

This vignette demonstrates complete end-to-end examples using only exported helpers: `create_table()`, `create_figure()`, `create_text()`, `define_cols()`, `add_stub_column()`, `add_style()`, `f_combine()`, `create_report()`, `save_report()`, and session helpers `tfl_set_options()` / `tfl_get_options()`. When running locally, set the top chunk `eval=TRUE` to generate the sample data and plot, then execute examples in order.
