---
title: "Reporting Examples — ksTFL (v3)"
author: "ksTFL Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reporting Examples — ksTFL (v3)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette presents a realistic progression of reporting examples — from minimal tables/listings/figures/text to a full clinical-style table with multi-level headers, spanning stub columns, multi-line titles/footnotes, reusable styles, and session defaults. All examples use exported ksTFL helpers only; set `eval=TRUE` locally to run the R chunks.
## Workflow overview

This vignette follows a consistent end-to-end pattern for each example so you can reproduce the full reporting pipeline locally:

- Prepare small, self-contained sample data (run the setup chunk with `eval=TRUE`).
- Build `TFL_spec` objects using `create_table()`, `create_figure()` or `create_text()`.
- Tune columns and styles via `define_cols()` and `add_style()` (use `f_combine()` to reference combined styles).
- Assemble specs into a `TFL_report` with `create_report()` (this consolidates style references and assigns `dataRef`).
- Export artifacts for rendering with `save_report()` (writes JSON and data/figure files into `metaPath`).

Best practices:

- Use named styles (`add_style()`) and reference them by id; avoid inspecting or mutating internal fields.
- Use `print(spec)` and `print(report)` for concise, user-facing previews rather than reading internals.
- Use `metaPath = tempdir()` during examples to avoid writing permanent files while you experiment.

```{r create_sample_data, eval = FALSE}
# Sample datasets used by multiple sections (run locally before executing examples)
set.seed(2025)

demog_tbl <- data.frame(
  subject_id = sprintf("S%03d", 1:24),
  age = sample(25:80, 24, TRUE),
  sex = sample(c("M","F"), 24, TRUE),
  trt = sample(c("Placebo","DrugA"), 24, TRUE),
  stringsAsFactors = FALSE
)

vitals_tbl <- data.frame(
  subject_id = rep(demog_tbl$subject_id, each = 2),
  visit = rep(c("Baseline","Week 12"), times = 24),
  sbp = round(rnorm(48, 120, 12)),
  dbp = round(rnorm(48, 75, 8)),
  stringsAsFactors = FALSE
)

labs_tbl <- data.frame(
  subject_id = demog_tbl$subject_id,
  ALT = round(rnorm(24, 28, 9)),
  AST = round(rnorm(24, 26, 8)),
  stringsAsFactors = FALSE
)

# Example plot file (created locally when running the vignette)
plot_file <- file.path(tempdir(), "example_plot.png")
png(plot_file, width = 600, height = 400)
plot(demog_tbl$age, vitals_tbl$sbp[1:24], xlab = "Age", ylab = "SBP", main = "Age vs SBP")
dev.off()
```

## 1 — Simple minimal table

Goal: show a single-table workflow from data to a `TFL_spec` using only exported helpers.

```{r example_minimal_table, eval = FALSE}
spec_min_table <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))
spec_min_table <- add_title(spec_min_table, "Demographics (minimal)")

# Print shows a compact overview (useful in interactive sessions)
print(spec_min_table)
# close the example chunk
```

```{r save_min_table, eval = FALSE}
# End-to-end: wrap single spec into a report and save
rpt_min_table <- create_report(spec_min_table)
res_min_table <- save_report(rpt_min_table, docFileName = "tbl_min.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_min_table
```

Notes: `print()` on a `TFL_spec` provides a readable overview including docType, titles and defined columns.

---

## 2 — Simple minimal figure

Goal: create a figure spec from a local image file and add a title.

```{r example_minimal_figure, eval = FALSE}
spec_min_fig <- create_figure(plot_file)
spec_min_fig <- add_title(spec_min_fig, "Example: Age vs SBP")
print(spec_min_fig)
# close example chunk
```

```{r save_min_fig, eval = FALSE}
# End-to-end: save the single-figure report
rpt_min_fig <- create_report(spec_min_fig)
res_min_fig <- save_report(rpt_min_fig, docFileName = "fig_min.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_min_fig
```

---

## 3 — Simple minimal text (narrative)

Goal: create a text-only spec for narrative content.

```{r example_minimal_text, eval = FALSE}
spec_min_text <- create_text()
spec_min_text <- add_body_text(spec_min_text, "This narrative describes the study population and analysis approach.")
print(spec_min_text)
# close example chunk
```

```{r save_min_text, eval = FALSE}
# End-to-end: save the narrative as a report
rpt_min_text <- create_report(spec_min_text)
res_min_text <- save_report(rpt_min_text, docFileName = "txt_min.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_min_text
```

---

## 4 — Define columns: single, batch, and parameter recycling

Goal: demonstrate `define_cols()` for both single-column and batch updates with flexible parameter mapping.

### Why `define_cols()`?
After creating a table with `create_table()`, column properties are auto-detected from the data. However, you often need to:
- Customize column labels for readability
- Specify numeric formats (e.g., "%.2f" for 2 decimal places)
- Control visibility, styling, or special behaviors (ID column, deduplicate, page breaks)
- Lock or adjust column widths for better layout

`define_cols()` is the primary tool for these customizations. It supports both **single-column** and **batch updates**, with intelligent **parameter recycling** to keep code concise.

### Example 1: Single column definition

The simplest case — modify one column at a time:

```{r example_define_single, eval = FALSE}
# Define one column
spec_single <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))
spec_single <- define_cols(spec_single, subject_id, label = "Subject ID", isID = TRUE)

# Another single column call (chaining is encouraged)
spec_single <- define_cols(spec_single, age, 
                           label = "Age (years)", 
                           type = "numeric", 
                           format = "%.0f")

print(spec_single)
# close example chunk
```

### Example 2: Batch update with single value (recycling to all columns)

Update multiple columns with the **same value** — the value is automatically recycled:

```{r example_define_batch_single, eval = FALSE}
# Apply single format to multiple columns
spec_batch <- create_table(data = demog_tbl, cols = c(age, height, weight))
spec_batch <- define_cols(spec_batch, c(age, height, weight),
                          type = "numeric",      # Applied to all 3
                          format = "%.1f",       # Applied to all 3
                          valueStyleRef = "numeric_right")  # Applied to all 3

print(spec_batch)
# close example chunk
```

**Why this matters**: Instead of writing three separate `define_cols()` calls, you write one line and the package applies the same values to all selected columns.

### Example 3: Batch update with per-column values (1-to-N mapping)

Update multiple columns with **different values** — provide a vector matching the number of columns:

```{r example_define_batch_mapped, eval = FALSE}
# Different label and format for each column
spec_mapped <- create_table(data = demog_tbl, cols = c(age, sex, trt))
spec_mapped <- define_cols(spec_mapped, c(age, sex, trt),
                           label = c("Age (years)", "Biological Sex", "Treatment Group"),
                           type = c("numeric", "string", "string"),
                           format = c("%.0f", "", "")
)

print(spec_mapped)
# close example chunk
```

**Important**: The vector length must match exactly:
- Length 1: applies to all columns
- Length N (where N = number of columns): applies one-to-one
- Any other length: raises an error

### Example 4: Multiple `define_cols()` calls (chaining with `|>`)

Combine `define_cols()` calls to layer customizations — each call merges with previous settings (last-win strategy):

```{r example_define_chain, eval = FALSE}
# Start with basic table
spec_chain <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))

# First pass: set all labels
spec_chain <- define_cols(spec_chain, c(subject_id, age, sex, trt),
                          label = c("Subject ID", "Age", "Sex", "Treatment"))

# Second pass: set numeric formatting for age and trt
spec_chain <- define_cols(spec_chain, age, type = "numeric", format = "%.0f")

# Third pass: mark subject_id as ID column (repeats on page breaks)
spec_chain <- define_cols(spec_chain, subject_id, isID = TRUE)

print(spec_chain)
# close example chunk
```

This approach makes it easy to:
- Define labels first (human-readable column names)
- Then apply formatting (numeric/string types, decimals)
- Then apply special behavior flags (ID, deduplicate, etc.)

---

## 5 — Set document properties (hasData, content width, placement)

Goal: configure document-level properties like data availability flag, content width, and placement.

**Key function**: `set_document()`
- `glueNumType`: Whether to combine type and number with first title (default TRUE)
- `hasData`: Whether document contains data (important for Text specs)
- `bodyTitles`, `bodyFootnotes`, `bodySubtitles`: Place these in body vs header
- `contentWidth`: Content width (e.g., "100%", "6.5in", "16.51cm")
- `isContinues`: Ignore page breaks between sections

**Example 1: hasData flag**

```{r example_set_document_prefix, eval = FALSE}
spec <- create_table(demog_tbl)

spec <- set_document(spec,
  hasData = TRUE)

print(spec)
```

**Example 2: Content width and placement**

```{r example_set_document_width, eval = FALSE}
spec <- create_table(demog_tbl)

spec <- set_document(spec,
  contentWidth = "95%",        # Narrower content (default 100%)
  bodyTitles = FALSE,          # Titles in header, not body
  bodyFootnotes = TRUE,        # Footnotes in body
  glueNumType = FALSE)
```

**Notes**:
- Most defaults are sensible; you typically only need `set_document()` for content width
- Multiple calls merge with last-win strategy (later calls override earlier ones)
- Pairs well with `set_page_style()` for complete document configuration

---

## 6 — Combine table/figure/text into a single report

Goal: assemble mixed specs into a `TFL_report` and inspect its summary.

```{r example_combine_report, eval = FALSE}
report_simple <- create_report(spec_min_table, spec_min_fig, spec_min_text)
print(report_simple)
# close example chunk
```

```{r save_report_simple, eval = FALSE}
# Save combined simple report
res_report_simple <- save_report(report_simple, docFileName = "report_simple.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_report_simple
```

Notes: `create_report()` preserves input order and consolidates styles and ids.

---

## 7 — Column widths: automatic calculation and locking

Goal: understand how column widths are automatically calculated and how to lock specific columns while others auto-adjust.

### How automatic column width works

When you create a table with `create_table()`, ksTFL automatically analyzes the data and distributes column widths proportionally:

1. **Initial analysis**: For each column, the package examines data values (length, type, number of decimals)
2. **Width calculation**: Numeric columns typically get wider if they have many decimals or large values; short text columns get narrower
3. **Proportional distribution**: All widths are normalized to sum to 100% (or remaining available width)
4. **Default**: `autoColWidth = TRUE` in package options (can be changed with `tfl_set_options()`)

**Example**: A table with columns `id` (short numeric), `description` (long text), `value` (numeric):
- Auto-detected widths might be: `id = 20%`, `description = 50%`, `value = 30%`

### Example 1: Accept auto-calculated widths

The simplest approach — let the package handle width distribution:

```{r example_colwidth_auto, eval = FALSE}
# Create table; widths are auto-calculated from data characteristics
spec_auto <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))
spec_auto <- define_cols(spec_auto, c(subject_id, ALT, AST),
                         label = c("Subject ID", "ALT (U/L)", "AST (U/L)"))

# Print to see auto-calculated widths
print(spec_auto)
# close example chunk
```

### Example 2: Lock one column, auto-adjust others

Lock a specific column width while others recalculate to fill remaining space:

```{r example_colwidth_lock_one, eval = FALSE}
# Lock subject_id at 15%, let ALT and AST split the remaining 85%
spec_lock1 <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))
spec_lock1 <- define_cols(spec_lock1, subject_id, 
                          label = "Subject ID",
                          colWidth = "15%")  # Lock at 15%

# When auto-recalculation runs (automatic), ALT and AST widths are recalculated
# to maintain their initial proportion while filling the remaining 85%

print(spec_lock1)
# close example chunk
```

### Example 3: Lock multiple columns with relative widths

Lock several columns and let others auto-adjust:

```{r example_colwidth_lock_multiple, eval = FALSE}
# Lock two columns, let the third auto-adjust
spec_lock_multi <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))
spec_lock_multi <- define_cols(spec_lock_multi, subject_id, 
                               label = "Subject ID",
                               colWidth = "12%")
spec_lock_multi <- define_cols(spec_lock_multi, ALT,
                               label = "ALT (U/L)",
                               colWidth = "40%")

# AST automatically fills remaining 48% (100% - 12% - 40%)

print(spec_lock_multi)
# close example chunk
```

### Example 4: Mixed units (percentages and absolute)

Combine percentage-based widths with absolute units:

```{r example_colwidth_mixed_units, eval = FALSE}
# Lock subject_id at 2 cm, others in percentages
spec_mixed <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))
spec_mixed <- define_cols(spec_mixed, subject_id,
                          label = "Subject ID",
                          colWidth = "2cm")  # Absolute width
spec_mixed <- define_cols(spec_mixed, ALT,
                          label = "ALT (U/L)",
                          colWidth = "35%")   # Percentage of remaining

print(spec_mixed)
# close example chunk
```

**Important**: When mixing units (cm, in, pt) with percentages, the absolute widths are reserved first, then percentages are calculated from remaining space.

### Example 5: Width validation and constraints

The package validates column widths to prevent invalid configurations:

```{r example_colwidth_validation, eval = FALSE}
spec_valid <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))

# VALID: Set a reasonable relative width
spec_valid <- define_cols(spec_valid, ALT,
                          colWidth = "30%")  # OK: 30% is valid

# INVALID: Cannot set 100% relative width (leaves no space for other columns)
# This will raise an error:
# spec_valid <- define_cols(spec_valid, ALT, colWidth = "100%")
# Error: relative width 100.0% exceeds maximum allowed 75.0% 
#        (accounting for locked columns and minimum 0.5% for other columns)

# INVALID: Cannot set width below minimum threshold
# spec_valid <- define_cols(spec_valid, AST, colWidth = "0.2%")
# Error: relative width 0.2% is below minimum 0.5%

# Minimum width constraints (can be customized via package options):
# - Relative widths (%): minimum 0.5%
# - Fixed widths (cm, in, pt): minimum 0.2cm (~2mm, ~0.08in)

# Configure minimum width via package options:
tfl_set_options(minColWidth = 1.0)  # Set minimum relative width to 1.0%

# Now 0.5% will fail, but 1.0% will succeed
spec_valid <- define_cols(spec_valid, AST,
                          colWidth = "1.0%")  # OK with new minimum

print(spec_valid)
```

**Width validation rules**:
- **Relative width (%)**: Must be ≥ `minColWidth` (default 0.5%)
- **Relative width (%)**: Cannot exceed max allowed considering locked columns and other column minimums
- **Fixed width**: Must be ≥ 0.2cm (all units: cm, in, pt, mm automatically converted)
- **Incompatible widths**: Cannot set width to 100% (would exclude all other columns)

**Error messages** are detailed and show the maximum allowed width when you exceed limits:
```
Error: relative width 100.0% exceeds maximum allowed 75.0% 
       (accounting for 2 locked columns requiring 25% total, 
        and minimum 0.5% for remaining 1 unlocked column)
```

---

## 8 — Table with multiple levels (headers/subheaders), titles and footnotes

Goal: build a clinical-style table with multi-level column headers, subtitles and footnotes.

```{r example_multilevel_table, eval = FALSE}
spec_multi <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))
spec_multi <- add_title(spec_multi, "Laboratory Results")
spec_multi <- add_subtitle(spec_multi, "Selected hepatic enzymes by subject")
spec_multi <- add_footnote(spec_multi, "Values are shown as observed.")

spec_multi <- define_cols(
  spec_multi,
  c(subject_id, ALT, AST),
  label = c("Subject ID", "ALT (U/L)", "AST (U/L)")
)

print(spec_multi)
# close example chunk
```

```{r save_multilevel, eval = FALSE}
# Save the multilevel table as a report
rpt_multi <- create_report(spec_multi)
res_multi <- save_report(rpt_multi, docFileName = "tbl_multi.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_multi
```

---

## 9 — Stub columns (spanning headers)

Goal: demonstrate `add_span_header()` to create stacked/spanning stub headers above groups of columns.

### Why spanning headers?

In clinical tables, you often need to group related columns under a common header. For example:
- "Baseline" spanning columns: `visit`, `sbp`, `dbp`, `pulse`
- "Week 12" spanning columns: `sbp`, `dbp`, `pulse` (repeated measurements at different visits)

Spanning headers (called "stubs" in ksTFL) are separate from column labels — they sit above the column labels and group multiple columns. Multiple stubs can be stacked at different vertical levels.

### Example 1: Single spanning header

Create one stub that groups related columns:

```{r example_stubs_single, eval = FALSE}
# Start with demographics table
spec_stub_simple <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))
spec_stub_simple <- define_cols(spec_stub_simple, 
                                c(subject_id, age, sex, trt), 
                                label = c("Subject","Age","Sex","Treatment"))

# Add a spanning header for "Demographics" above age and sex
spec_stub_simple <- add_span_header(spec_stub_simple, 
                                    cols = c("age", "sex"), 
                                    label = "Demographics")

print(spec_stub_simple)
# close example chunk
```

**Result**: A table with column labels on one row and a "Demographics" header spanning age+sex columns above it.

### Example 1b: Using tidyselect helpers with spanning headers

`add_span_header()` supports all tidyselect expressions — use helpers for flexible column selection:

```{r example_stubs_tidyselect, eval = FALSE}
# Table with mixed column types
mixed_data <- data.frame(
  id = 1:10,
  age_baseline = rnorm(10, 45, 10),
  age_follow = rnorm(10, 46, 10),
  weight_baseline = rnorm(10, 70, 10),
  weight_follow = rnorm(10, 71, 10)
)

spec_tidysel <- create_table(mixed_data)

# Using starts_with() helper
spec_tidysel <- add_span_header(spec_tidysel, 
                                cols = starts_with("age"),
                                label = "Age Measurements",
                                stubOrder = 0)

# Using negation (-) to exclude columns
spec_tidysel <- add_span_header(spec_tidysel,
                                cols = -id,
                                label = "Baseline and Follow",
                                stubOrder = 1)

# Using matches() regex pattern
spec_tidysel <- add_span_header(spec_tidysel,
                                cols = matches("_baseline$"),
                                label = "Baseline Visits",
                                stubOrder = 2)

print(spec_tidysel)
```

**Tidyselect expressions supported**:
- **Ranges**: `cols = age:weight` (all columns between age and weight)
- **Helpers**: `cols = starts_with("age")`, `contains("baseline")`, `matches("^w")`
- **Negation**: `cols = -id` (all columns except id)
- **Combinations**: `cols = c(starts_with("age"), weight_baseline)`

### Example 2: Multiple stubs at different levels

Stack multiple stubs to create a multi-level hierarchy:

```{r example_stubs_multi, eval = FALSE}
# Create table with visit measurements at two time points
vitals_data <- data.frame(
  subject = sprintf("S%03d", 1:10),
  baseline_sbp = round(rnorm(10, 120, 10)),
  baseline_dbp = round(rnorm(10, 75, 8)),
  week12_sbp = round(rnorm(10, 118, 10)),
  week12_dbp = round(rnorm(10, 74, 8))
)

# Create spec and define columns
spec_stubs_multi <- create_table(data = vitals_data, 
                                 cols = c(subject, baseline_sbp, baseline_dbp, week12_sbp, week12_dbp))
spec_stubs_multi <- define_cols(spec_stubs_multi,
                                c(subject, baseline_sbp, baseline_dbp, week12_sbp, week12_dbp),
                                label = c("Subject ID", "SBP", "DBP", "SBP", "DBP"))

# Add first stub spanning baseline measurements
spec_stubs_multi <- add_span_header(spec_stubs_multi,
                                    cols = c("baseline_sbp", "baseline_dbp"),
                                    label = "Baseline",
                                    stubOrder = 1)

# Add second stub spanning week 12 measurements
spec_stubs_multi <- add_span_header(spec_stubs_multi,
                                    cols = c("week12_sbp", "week12_dbp"),
                                    label = "Week 12",
                                    stubOrder = 2)

print(spec_stubs_multi)
# close example chunk
```

**How `stubOrder` works**: 
- Lower numbers appear closer to the top (e.g., `stubOrder = 1` above `stubOrder = 2`)
- You can have multiple stubs at the same order level as long as their column sets don't overlap
- If NULL, `stubOrder` is auto-generated starting from 1

### Example 3: Styled spanning headers

Apply styles to stub labels using `labelStyleRef`:

```{r example_stubs_styled, eval = FALSE}
# First, create a style for stub labels
spec_stubs_style <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))

spec_stubs_style <- add_style(spec_stubs_style, 
                              id = "stub_header",
                              s_font(bold = TRUE, font_size = "12pt"),
                              s_paragraph(alignment = "center"),
                              s_table_style(background_color = "#E8E8E8"))

# Define columns
spec_stubs_style <- define_cols(spec_stubs_style,
                                c(subject_id, age, sex, trt),
                                label = c("Subject ID", "Age", "Sex", "Treatment"))

# Add stub with style reference
spec_stubs_style <- add_span_header(spec_stubs_style,
                                    cols = c("age", "sex"),
                                    label = "Demographics",
                                    labelStyleRef = "stub_header")

print(spec_stubs_style)
# close example chunk
```

Notes: 
- `labelStyleRef` can be a single style name or multiple styles combined with `f_combine()`
- Stub labels inherit the applied style, making grouped columns visually distinct
- Styles must be defined before referencing them in `add_span_header()`

```{r save_stub, eval = FALSE}
# Save stubbed table report
rpt_stub <- create_report(spec_stubs_style)
res_stub <- save_report(rpt_stub, docFileName = "tbl_stub.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_stub
```

---

## 10 — Styles and `f_combine()` for combined style references

Goal: create reusable named styles and combine them flexibly using `f_combine()`.

### Why named styles?

Styles are foundational in professional reporting:
- Define once, reference many times — consistency across your document
- Easy to update: change one style definition and all references automatically pick up the change
- Composable: combine base styles (bold, red text) into complex styles for specific use cases

ksTFL uses a **named style system**: you define styles with `add_style()` giving each an `id`, then reference them by name wherever you need them (`labelStyleRef`, `valueStyleRef`, `labelStyleRef` in stubs, etc.).

### Example 1: Define base styles

Create atomic styles that focus on one aspect (font, alignment, color):

```{r example_styles_base, eval = FALSE}
# Create a table spec
spec_base_styles <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))

# Define reusable base styles
spec_base_styles <- add_style(spec_base_styles, id = "bold_header",
                              s_font(bold = TRUE, font_size = "12pt"))

spec_base_styles <- add_style(spec_base_styles, id = "right_align",
                              s_paragraph(alignment = "right"))

spec_base_styles <- add_style(spec_base_styles, id = "light_gray_bg",
                              s_table_style(background_color = "#F5F5F5"))

# Apply to columns
spec_base_styles <- define_cols(spec_base_styles,
                                c(subject_id, age, sex, trt),
                                label = c("Subject ID", "Age", "Sex", "Treatment"),
                                labelStyleRef = c("bold_header", "", "", ""))

print(spec_base_styles)
# close example chunk
```

### Example 2: Combine styles with `f_combine()`

Apply multiple styles to a single element using `f_combine()` — they merge at render time:

```{r example_styles_fcombine, eval = FALSE}
# Create spec and define base styles
spec_combined <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))

# Define atomic styles
spec_combined <- add_style(spec_combined, id = "bold_text", s_font(bold = TRUE))
spec_combined <- add_style(spec_combined, id = "red_color", s_font(color = "#CC0000"))
spec_combined <- add_style(spec_combined, id = "centered", s_paragraph(alignment = "center"))

# Combine styles: bold + red text + centered
spec_combined <- define_cols(spec_combined,
                             subject_id,
                             label = "Subject ID",
                             labelStyleRef = f_combine("bold_text", "red_color", "centered"))

# Different columns can use different combinations
spec_combined <- define_cols(spec_combined, age, label = "Age",
                             labelStyleRef = f_combine("bold_text", "centered"))

print(spec_combined)
# close example chunk
```

**How `f_combine()` works**:
- Takes multiple style names as arguments
- Returns a reference object that tells the package to merge those styles
- Useful for creating ad-hoc combinations without defining new named styles
- Order matters for last-win conflict resolution (later arguments override earlier ones)

### Example 3: Batch apply combined styles to multiple columns

Use the recycling feature to apply the same combined style to multiple columns:

```{r example_styles_batch_combine, eval = FALSE}
# Create spec and styles
spec_batch_combined <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))

spec_batch_combined <- add_style(spec_batch_combined, id = "header_emphasis",
                                 s_font(bold = TRUE, font_size = "13pt"),
                                 s_table_style(background_color = "#E0E0E0"))

# Apply combined styles: emphasis for subject_id + age, different for sex + trt
spec_batch_combined <- define_cols(spec_batch_combined,
                                   c(subject_id, age, sex, trt),
                                   label = c("Subject ID", "Age", "Sex", "Treatment"),
                                   labelStyleRef = c(f_combine("header_emphasis", "centered"),
                                                     f_combine("header_emphasis", "centered"),
                                                     "centered",
                                                     "centered"))

print(spec_batch_combined)
# close example chunk
```

**Key insight**: When using multiple columns with different style combinations, provide a vector matching the column count (one-to-one mapping).

```{r save_styles, eval = FALSE}
# Save the styled table report
rpt_styles <- create_report(spec_batch_combined)
res_styles <- save_report(rpt_styles, docFileName = "tbl_styles.docx", outDir = "./out", metaPath = tempdir(), prettify = TRUE)
res_styles
```

---

## 11 — Conditional row actions with `compute_cols()`

Goal: demonstrate conditional styling, merging, and row insertion using `compute_cols()`.

**Overview**: While `define_cols()` sets properties globally for all rows, `compute_cols()` applies **conditional actions** to specific rows matching a condition. Common use cases:
- Style rows where a specific value occurs (e.g., first/last occurrence, threshold-based)
- Merge columns in certain rows (e.g., group headers)
- Insert separator or summary rows programmatically

### Example 1: Conditional styling

Apply a style to columns in rows matching a condition:

```{r compute_basic_style, eval = FALSE}
# Sample data with groups
data <- data.frame(
  group = c("A", "A", "B", "B", "C"),
  metric = c("Value 1", "Value 2", "Value 3", "Value 4", "Value 5"),
  count = c(10, 20, 15, 25, 30)
)

# Create spec and define styles
spec <- create_table(data) |>
  add_style("group_header", s_font(bold = TRUE, color = "#0000FF")) |>
  add_style("emphasize", s_table_style(background_color = "#FFFFCC"))

# Apply conditional styling: bold+blue for first occurrence of each group
spec <- spec |>
  compute_cols(firstOf(group), c_style(c(metric, count), styleRef = "group_header"))

# Apply conditional styling: highlight rows with high count
spec <- spec |>
  compute_cols(count > 20, c_style(count, styleRef = "emphasize"))

print(spec)
```

### Example 2: Combining styles in conditional rows

Apply multiple styles to a single column using `f_combine()`:

```{r compute_combined_style, eval = FALSE}
spec <- create_table(data) |>
  add_style("bold", s_font(bold = TRUE)) |>
  add_style("large", s_font(font_size = "14pt")) |>
  add_style("highlight", s_table_style(background_color = "#FFFF00"))

# Apply combined styles to first group occurrence
spec <- spec |>
  compute_cols(firstOf(group), 
    c_style(metric, styleRef = f_combine("bold", "large", "highlight"))
  )
```

### Example 3: Column merging in conditional rows

Merge adjacent columns for rows matching a condition:

```{r compute_merge, eval = FALSE}
spec <- create_table(data) |>
  add_style("group_label", s_table_style(background_color = "#D9D9D9"))

# Merge metric and count columns for first occurrence (group header style)
spec <- spec |>
  compute_cols(firstOf(group), 
    c_merge(c(metric, count), styleRef = "group_label")
  )
```

### Example 4: Inserting separator rows

Insert empty rows (separators) or rows with content from another column:

```{r compute_addrow, eval = FALSE}
spec <- create_table(data) |>
  add_style("separator", s_table_style(background_color = "#E8E8E8"))

# Insert empty separator above first group occurrence
spec <- spec |>
  compute_cols(firstOf(group), 
    c_addrow(pos = "above")  # Empty row, no value_from
  )

# Insert summary row below last group occurrence (using a specific column as content)
spec <- spec |>
  compute_cols(lastOf(group), 
    c_addrow(pos = "below", value_from = "group", styleRef = "separator")
  )
```

### Example 5: Multiple actions on same rows

Combine styling, merging, and row insertion in a single `compute_cols()` call:

```{r compute_multi_action, eval = FALSE}
spec <- create_table(data) |>
  add_style("header", s_font(bold = TRUE)) |>
  add_style("separator", s_table_style(background_color = "#E8E8E8"))

# For first group occurrence: add separator, style, and merge
spec <- spec |>
  compute_cols(firstOf(group), 
    c_addrow(pos = "above"),  # Add empty separator
    c_style(metric, styleRef = "header"),  # Style metric column
    c_merge(c(metric, count))  # Merge adjacent columns
  )
```

**Key concepts**:
- **Conditions** are unevaluated expressions evaluated at report generation time (during `create_report()`)
- **Multiple calls accumulate**: calling `compute_cols()` multiple times on the same spec appends actions
- **Multiple actions in one call**: same row can have styling, merging, and row insertion simultaneously
- **Overlapping conditions**: if multiple conditions match the same row, all actions apply (styling aggregates, merging rules apply)
- **value_from**: Optional in `c_addrow()` — omit for empty separator rows
- **Performance**: Conditions evaluated once per row during report assembly; style consolidation happens automatically

---

## 12 — Save report (end-to-end export)

Goal: show `save_report()` usage — it validates and serializes specs and writes metadata/data/figure files to `metaPath`.

### What `save_report()` does

`save_report()` is the final step in the reporting pipeline. It:
1. **Assembles** multiple specs into a single report (if not already done via `create_report()`)
2. **Validates** the complete report against JSON Schema
3. **Serializes** to JSON, replacing internal references with file paths
4. **Writes files** to `metaPath`:
   - `{spec_hash}.json` — the main metadata spec file (consumed by the renderer)
   - `{dataRef}.json` — table data files (one per table spec)
   - Figure files — copied from input paths to `metaPath`
5. **Returns metadata** with generation timestamps and file locations

### Example: Assemble and save a multi-spec report

```{r example_save_report, eval = FALSE}
# Create multiple specs
table_spec <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))
table_spec <- add_title(table_spec, "Laboratory Results")

text_spec <- create_text()
text_spec <- add_body_text("All values are from the locked database.")

# Assemble into report
report_full <- create_report(table_spec, text_spec)

# Save — internally validates and writes JSON + data files to metaPath
res <- save_report(report_full, 
                   docFileName = "example_report.docx", 
                   outDir = "./out", 
                   metaPath = tempdir(), 
                   prettify = TRUE)

# Inspect metadata
str(res)
```

**Parameters**:
- `report`: A `TFL_report` object (created via `create_report()`)
- `docFileName`: Output DOCX filename (passed to renderer)
- `outDir`: Directory where output DOCX will be written
- `metaPath`: Directory where metadata and data files are written (renderer consumes these)
- `prettify`: If TRUE, formats JSON output nicely (useful for debugging)

**Return value**: A list with:
- `spec_file`: Full path to generated spec JSON
- `data_files`: Named list of generated data file paths
- `metadata`: Report-level metadata (timestamps, paths)

Notes: `save_report()` is the public entrypoint to persist specs/data/figures for downstream rendering. In a real workflow, `render_docx()` consumes the spec JSON to produce the final .docx file.

---

## 13 — Session-wide `tfl_options`: common headers/footers and default body text

Goal: set session defaults so multiple specs inherit common headers, footers, and default body text without repeating code.

### Why session options?

In a real clinical reporting workflow, many tables/listings share:
- Common headers (study name, database version, confidentiality notice)
- Common footers (company name, page numbers, disclaimers)
- Default body text (standard disclaimers or methodology notes)

Instead of adding these to every spec, use `tfl_set_options()` to set session defaults once. All specs created afterward inherit these settings.

### Example 1: Set basic session options

```{r example_session_options_basic, eval = FALSE}
# Set session defaults (applies to all NEW specs created after this call)
tfl_set_options(
  add_header(c("Study ABC", "Phase II Safety Study", "CONFIDENTIAL")),
  add_footer(c("Company Confidential", "Page {page} of {numpages}"))
)

# Create spec — automatically inherits headers/footers from options
spec_with_opts <- create_table(data = demog_tbl, cols = c(subject_id, age, sex, trt))
spec_with_opts <- add_title(spec_with_opts, "Demographics Table")

print(spec_with_opts)
# close example chunk
```

**Key point**: Headers and footers from `tfl_set_options()` are automatically applied to new specs. No need to call `add_header()` again.

### Example 2: Override session options in a specific spec

You can override session defaults on individual specs:

```{r example_session_options_override, eval = FALSE}
# Session options are still in effect from previous example

# Create a spec with session defaults
spec_default <- create_table(data = labs_tbl, cols = c(subject_id, ALT, AST))
spec_default <- add_title(spec_default, "Lab Results (using session defaults)")

# Create another spec but override the header
spec_override <- create_table(data = demog_tbl, cols = c(subject_id, age))
spec_override <- add_header(c("Study XYZ", "Different Study", "CONFIDENTIAL"))  # Overrides session default
spec_override <- add_title(spec_override, "Demographics (custom header)")

print(spec_default)   # Uses session header from tfl_set_options
print(spec_override)  # Uses custom header from add_header call
# close example chunk
```

**Rule**: Spec-level settings always take precedence over session defaults (last-win strategy).

### Example 3: Check and reset session options

Inspect current session options and reset to defaults:

```{r example_session_options_inspect, eval = FALSE}
# Check current options
current_options <- tfl_get_options()
str(current_options)

# Get a single option
current_missings <- tfl_get_option("missings")
cat("Current missings representation:", current_missings, "\n")

# Reset to package defaults
tfl_reset_options()

# Now new specs will use package defaults instead of your custom session options
spec_reset <- create_table(data = demog_tbl, cols = c(subject_id, age))
print(spec_reset)
# close example chunk
```

---

## Final notes

This comprehensive vignette demonstrates the complete ksTFL reporting pipeline using only exported helpers. 

**Examples covered**:
1. Minimal table/figure/text creation (single-spec workflows)
2. **Define columns**: Single column, batch updates, parameter recycling (NEW)
3. Column widths: Auto-calculation, locking, mixed units (NEW)
4. Multi-level tables with titles, subtitles, and footnotes
5. Combining multiple specs into a report
6. Spanning headers (stub columns) with styling
7. Named styles and combining with `f_combine()` (expanded)
8. End-to-end export with `save_report()`
9. Session defaults with `tfl_set_options()`

**Key patterns to remember**:
- **Batch operations**: Use `c()` to select multiple columns, provide single values (recycled) or N values (one-to-one mapping)
- **Multiple calls**: Chain `define_cols()`, `add_style()`, etc. — each call merges with previous settings
- **Flexible column widths**: Auto-calculated by default; lock specific columns with `colWidth` to trigger recalculation of others
- **Combining styles**: Use `f_combine()` for on-the-fly combinations; define named styles for reuse
- **Session defaults**: Use `tfl_set_options()` once; all new specs inherit settings (override per-spec if needed)

**Running examples locally**: Set the top chunk `eval=TRUE` to generate sample data, then execute examples in order. All code uses exported functions only — no internal API manipulation needed.

For more details on function parameters, see the roxygen documentation: `?create_table`, `?define_cols`, `?add_style`, `?save_report`, etc.

