---
title: "Styling Guide — ksTFL"
author: "ksTFL Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Styling Guide — ksTFL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", collapse = TRUE, eval = FALSE)
```

## Overview

This guide covers the complete styling system in ksTFL:
- **Style primitives** (`s_*` helpers) for fonts, paragraphs, spacing, indentation, tables, and borders
- **Declaring named styles** with `add_style()`
- **Referencing and combining styles** with style references and `f_combine()`
- **Applying styles** to columns, labels, stubs, and content
- **Best practices** for maintainable, reusable style systems

For runnable reporting examples integrating styles see [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd). For a quick start see [Getting Started](Getting_Started_with_ksTFL.Rmd).

---

## Styling philosophy and workflow

### Why declarative styles?

Professional clinical documents require:
- **Consistency**: All headers look the same, all numeric columns align right, all subtitles use the same font
- **Maintainability**: Change one style definition, all references update automatically
- **Composability**: Build complex styles from simple building blocks (font + paragraph alignment + table background)
- **Separation of concerns**: Define styles once, apply them many times without repeating details

ksTFL uses a **named style system**: you declare styles with `add_style()` giving each an `id`, then reference them by name where needed. Styles are composable (`f_combine()`) and consolidated automatically when you build reports.

### Best practices workflow

1. **Define atomic styles first**: Create small, focused styles (bold headers, right-aligned text, light gray background)
2. **Reference by name**: In `define_cols()`, `add_title()`, `add_span_header()` (with tidyselect support), use `labelStyleRef` or `valueStyleRef` to point to styles by id
3. **Combine when needed**: Use `f_combine()` to merge multiple styles on-the-fly for ad-hoc combinations
4. **Let `create_report()` consolidate**: The report builder automatically merges combined styles so the renderer receives clean, consolidated styles
5. **Never inspect internal fields**: Don't manually look at `spec$attribs$styles` — let the package manage consolidation

---

## Style primitives (s_* helpers)

All style primitives begin with `s_` and must be used **only inside `add_style()`**. The helpers validate parameter names and values, raising informative errors if used incorrectly.

### `s_font()` — Font properties

Control font appearance (name, size, weight, color, decorations).

**Parameters**:
- `font_name`: Font family (e.g., "Arial", "Courier New", "Times New Roman", "Calibri")
- `font_size`: Size with units (e.g., "12pt", "11pt")
- `bold`: Logical (TRUE/FALSE)
- `italic`: Logical (TRUE/FALSE)
- `underline`: Logical (TRUE/FALSE)
- `color`: Color as hex code (e.g., "#000000", "#FF0000") or color name (e.g., "red", "black", "blue")
- `highlight`: Background highlight color as hex code or color name

**Example**:
```{r s_font_example, eval = FALSE}
spec <- create_table(mtcars)

# Bold, 12pt Arial, red text (using hex code)
spec <- add_style(spec, id = "header_bold_red",
  s_font(font_name = "Arial", font_size = "12pt", bold = TRUE, color = "#CC0000"))

# Underlined, 10pt monospace, blue text (using color name)
spec <- add_style(spec, id = "code_style",
  s_font(font_name = "Courier New", font_size = "10pt", underline = TRUE, color = "blue"))

# Yellow highlight with black text
spec <- add_style(spec, id = "highlighted",
  s_font(color = "black", highlight = "yellow"))
```

### `s_paragraph()` — Paragraph properties

Control text alignment, spacing before/after, line spacing, and indentation.

**Parameters**:
- `alignment`: Text alignment — "left", "right", "center", "justify", "distributed"
- `spacing`: Spacing before/after (use `s_spacing(...)`)
- `indents`: Left/right indentation, first-line indent (use `s_indents(...)`)

**Example**:
```{r s_paragraph_example, eval = FALSE}
spec <- create_table(mtcars)

# Centered text with 6pt spacing before/after
spec <- add_style(spec, id = "centered_spaced",
  s_paragraph(alignment = "center", 
              spacing = s_spacing(before = "6pt", after = "6pt")))

# Right-aligned with left indent
spec <- add_style(spec, id = "right_indented",
  s_paragraph(alignment = "right",
              indents = s_indents(left = "10pt")))
```

### `s_spacing()` — Line and paragraph spacing

Controls spacing **before** a paragraph, **after** a paragraph, and **between lines**.

**Parameters**:
- `before`: Space before paragraph (e.g., "6pt", "12pt")
- `after`: Space after paragraph (e.g., "6pt", "12pt")
- `line_spacing`: Line spacing multiplier (e.g., 1.0, 1.5, 2.0 for single/1.5-line/double spacing)

**Note**: `s_spacing()` is always used **inside `s_paragraph()`**, never standalone.

**Example**:
```{r s_spacing_example, eval = FALSE}
spec <- create_table(mtcars)

# Double-spaced with 12pt spacing after each paragraph
spec <- add_style(spec, id = "double_spaced",
  s_paragraph(spacing = s_spacing(after = "12pt", line_spacing = 2.0)))
```

### `s_indents()` — Indentation

Controls left/right margins and first-line indentation within a paragraph.

**Parameters**:
- `left`: Left indent (e.g., "10pt", "1cm")
- `right`: Right indent (e.g., "10pt")
- `first_line`: First-line indent (e.g., "20pt" for hanging indent)

**Note**: `s_indents()` is always used **inside `s_paragraph()`**, never standalone.

**Example**:
```{r s_indents_example, eval = FALSE}
spec <- create_table(mtcars)

# Hanging indent (first line outdented, rest indented)
spec <- add_style(spec, id = "hanging_indent",
  s_paragraph(indents = s_indents(left = "20pt", first_line = "-20pt")))

# Left and right margins with left indent
spec <- add_style(spec, id = "block_indent",
  s_paragraph(indents = s_indents(left = "30pt", right = "30pt")))
```

### `s_table_style()` — Table cell properties

Control cell background, row height, vertical alignment, text orientation, and borders.

**Parameters**:
- `background_color`: Cell background color (hex, e.g., "#E8E8E8")
- `row_height`: Height of table row (e.g., "25pt")
- `vertical_alignment`: "top", "center", "bottom"
- `text_orientation`: "horizontal", "vertical_90", "vertical_270"
- `borders`: Border specification (use `s_borders(...)`)

**Example**:
```{r s_table_style_example, eval = FALSE}
spec <- create_table(mtcars)

# Light gray background, centered vertically, fixed row height
spec <- add_style(spec, id = "header_cell",
  s_table_style(background_color = "#F2F2F2",
                row_height = "30pt",
                vertical_alignment = "center"))

# Vertically rotated text (90 degrees)
spec <- add_style(spec, id = "rotated_header",
  s_table_style(text_orientation = "vertical_90"))
```

### `s_borders()` — Border specifications

Define borders for all four sides of a cell. Each side takes `s_border()` with line style, width, and color.

**Parameters** (each side):
- `top`: Top border (use `s_border(...)`)
- `bottom`: Bottom border (use `s_border(...)`)
- `left`: Left border (use `s_border(...)`)
- `right`: Right border (use `s_border(...)`)

**Note**: `s_borders()` is always used **inside `s_table_style()`**, never standalone.

### `s_border()` — Individual border line

Defines a single border line with style, width, and color.

**Parameters**:
- `color`: Color as hex code (e.g., "#000000") or color name (e.g., "black", "red")
- `width`: Line width (e.g., "1pt", "2pt", "0.5pt")
- `line_style`: "single", "double", "dashed", "dotted", "thick", "none"

**Example**:  
```{r s_borders_example, eval = FALSE}
spec <- create_table(mtcars)

# All borders: thin single lines in dark gray
spec <- add_style(spec, id = "all_borders",
  s_table_style(
    borders = s_borders(
      top = s_border(color = "grey40", width = "1pt", line_style = "single"),
      bottom = s_border(color = "grey40", width = "1pt", line_style = "single"),
      left = s_border(color = "grey40", width = "1pt", line_style = "single"),
      right = s_border(color = "grey40", width = "1pt", line_style = "single")
    )
  )
)

# Heavy bottom border in dark color (using hex code)
spec <- add_style(spec, id = "bottom_border_heavy",
  s_table_style(
    borders = s_borders(
      bottom = s_border(color = "#333333", width = "2pt", line_style = "thick")
    )
  )
)
```

---

## Declaring named styles with `add_style()`

Named styles are the foundation of the ksTFL styling system. Each style has a unique `id` and contains one or more style primitives.

### Basic syntax

```{r add_style_basic, eval = FALSE}
spec <- add_style(spec, id = "style_name",
  s_font(...),
  s_paragraph(...),
  s_table_style(...)
)
```

**Parameters**:
- `spec`: A `TFL_spec` object
- `id`: Unique name for the style (e.g., "header_bold", "numeric_right")
- `...`: One or more style primitives (`s_font()`, `s_paragraph()`, `s_table_style()`, etc.)

### Example: Define a complete style

```{r add_style_complete, eval = FALSE}
spec <- create_table(mtcars)

# Comprehensive header style: bold white text on gray background, centered
spec <- add_style(spec, id = "table_header",
  s_font(font_name = "Arial", font_size = "12pt", bold = TRUE, color = "#FFFFFF"),
  s_paragraph(alignment = "center", spacing = s_spacing(before = "6pt", after = "6pt")),
  s_table_style(background_color = "#333333", row_height = "30pt", vertical_alignment = "center")
)

# Simple numeric style: right-aligned
spec <- add_style(spec, id = "numeric_right",
  s_paragraph(alignment = "right")
)

# ID/key style: bold
spec <- add_style(spec, id = "id_bold",
  s_font(bold = TRUE)
)
```

### Multiple calls merge with last-win strategy

Calling `add_style()` multiple times with the same `id` merges the styles (last-win semantics):

```{r add_style_merge, eval = FALSE}
spec <- create_table(mtcars)

# First call: defines font
spec <- add_style(spec, id = "emphasis", s_font(bold = TRUE))

# Second call: adds paragraph alignment; bold is preserved
spec <- add_style(spec, id = "emphasis", s_paragraph(alignment = "center"))

# Result: "emphasis" has both bold font AND center alignment
```

---

## Referencing and applying styles

Once you declare styles with `add_style()`, reference them by id in various places:

### 1. Column labels — `labelStyleRef` in `define_cols()`

Apply styles to **column header labels**:

```{r apply_label_style, eval = FALSE}
spec <- create_table(mtcars)

spec <- add_style(spec, id = "header_bold", s_font(bold = TRUE, font_size = "12pt"))

# Apply to single column
spec <- define_cols(spec, mpg, label = "MPG (miles/gallon)", labelStyleRef = "header_bold")

# Apply to multiple columns with recycling
spec <- define_cols(spec, c(hp, cyl), label = c("HP", "Cylinders"), labelStyleRef = "header_bold")

# Different styles for different columns
spec <- define_cols(spec, c(mpg, hp),
  label = c("MPG", "HP"),
  labelStyleRef = c("header_bold", "header_italic"))
```

### 2. Column values — `valueStyleRef` in `define_cols()`

Apply styles to **data values** in a column:

```{r apply_value_style, eval = FALSE}
spec <- create_table(mtcars)

spec <- add_style(spec, id = "numeric_right", s_paragraph(alignment = "right"))

# Right-align all numeric values in the mpg column
spec <- define_cols(spec, mpg, type = "numeric", valueStyleRef = "numeric_right")
```

### 3. Spanning headers — `labelStyleRef` in `add_span_header()`

Apply styles to **stub (spanning header) labels**:

```{r apply_stub_style, eval = FALSE}
spec <- create_table(mtcars)

spec <- add_style(spec, id = "stub_label",
  s_font(bold = TRUE, font_size = "11pt"),
  s_table_style(background_color = "#EFEFEF"))

spec <- add_span_header(spec, cols = c("mpg", "cyl"), label = "Engine",
  labelStyleRef = "stub_label")
```

### 4. Content — `styleRef` in `add_title()`, `add_footnote()`

Apply styles to titles, subtitles, footnotes:

```{r apply_content_style, eval = FALSE}
spec <- create_table(mtcars)

spec <- add_style(spec, id = "title_style",
  s_font(bold = TRUE, font_size = "14pt"),
  s_paragraph(alignment = "center"))

spec <- add_title(spec, "Motor Trends Analysis", styleRef = "title_style")
```

---

## Combining styles with `f_combine()`

`f_combine()` lets you apply multiple styles to a single element without pre-defining a combined style.

### Basic usage

```{r f_combine_basic, eval = FALSE}
spec <- create_table(mtcars)

spec <- add_style(spec, id = "bold", s_font(bold = TRUE))
spec <- add_style(spec, id = "red", s_font(color = "red"))  # Using color name
spec <- add_style(spec, id = "centered", s_paragraph(alignment = "center"))

# Apply bold + red + centered to a column header
spec <- define_cols(spec, mpg, label = "MPG",
  labelStyleRef = f_combine("bold", "red", "centered"))
```

### When to use `f_combine()` vs named styles

| Use Case | Approach |
|----------|----------|
| **Reusable style** (used in 3+ places) | Define a named style with `add_style()` |
| **One-off combination** (used once or twice) | Use `f_combine()` inline |
| **Complex style** (many properties) | Define named style, then optionally combine with others |
| **Per-column variations** | Use `f_combine()` with per-column vectors |

### Combining with per-column mapping

Apply different combinations to different columns:

```{r f_combine_per_column, eval = FALSE}
spec <- create_table(mtcars)

# Define base styles
spec <- add_style(spec, id = "bold", s_font(bold = TRUE))
spec <- add_style(spec, id = "italic", s_font(italic = TRUE))
spec <- add_style(spec, id = "centered", s_paragraph(alignment = "center"))
spec <- add_style(spec, id = "right", s_paragraph(alignment = "right"))

# Apply different combinations per column
spec <- define_cols(spec, c(mpg, cyl, hp),
  label = c("MPG", "Cylinders", "HP"),
  labelStyleRef = c(
    f_combine("bold", "centered"),      # mpg: bold + centered
    f_combine("italic", "right"),       # cyl: italic + right
    f_combine("bold", "italic", "centered")  # hp: bold + italic + centered
  )
)
```

---

## Allowed values (enumerations)

### Color names
Color parameters accept both **hex codes** (e.g., `"#FF0000"`) and **named colors**:

**Basic colors**: black, white, red, green, blue, yellow, orange, purple, pink, brown, cyan, magenta, navy, teal, lime, maroon, olive, silver, gold, coral, salmon, turquoise, violet, indigo, khaki, lavender, plum, tan

**Grayscale**: grey10, grey20, grey30, grey40, grey60, grey70, grey80, grey90 (or gray with 'a')

### Font names (common)
- Arial
- Courier New
- Times New Roman
- Calibri
- Georgia
- Verdana

### Alignment
- "left"
- "right"
- "center"
- "justify"
- "distributed"

### Border line styles
- "single" (solid line)
- "double" (two lines)
- "dashed" (dashed line)
- "dotted" (dotted line)
- "thick" (thicker solid line)
- "none" (no border)

### Vertical alignment
- "top"
- "center"
- "bottom"

### Text orientation
- "horizontal" (default)
- "vertical_90" (rotated 90 degrees clockwise)
- "vertical_270" (rotated 270 degrees clockwise)

---

## Common styling patterns

### Pattern 1: Clinical table headers

```{r pattern_clinical_header, eval = FALSE}
spec <- create_table(my_data)

spec <- add_style(spec, id = "clinical_header",
  s_font(font_name = "Arial", font_size = "11pt", bold = TRUE, color = "#FFFFFF"),
  s_paragraph(alignment = "center"),
  s_table_style(background_color = "#003366", row_height = "28pt", vertical_alignment = "center")
)

spec <- define_cols(spec, c(all columns), labelStyleRef = "clinical_header")
```

### Pattern 2: Right-aligned numeric columns

```{r pattern_numeric, eval = FALSE}
spec <- add_style(spec, id = "numeric_format",
  s_font(font_name = "Courier New", font_size = "10pt"),
  s_paragraph(alignment = "right")
)

spec <- define_cols(spec, c(age, weight, dose), type = "numeric", valueStyleRef = "numeric_format")
```

### Pattern 3: ID/key columns (bold, wide)

```{r pattern_id_column, eval = FALSE}
spec <- add_style(spec, id = "id_column",
  s_font(bold = TRUE, font_size = "11pt"),
  s_paragraph(alignment = "left")
)

spec <- define_cols(spec, subject_id, 
  label = "Subject ID",
  labelStyleRef = "id_column",
  colWidth = "15%",
  isID = TRUE)
```

### Pattern 4: Multi-level headers with styled stubs

```{r pattern_multi_stub, eval = FALSE}
spec <- add_style(spec, id = "level1_stub",
  s_font(bold = TRUE, font_size = "12pt", color = "#FFFFFF"),
  s_table_style(background_color = "#666666", vertical_alignment = "center"))

spec <- add_style(spec, id = "level2_stub",
  s_font(bold = TRUE, font_size = "11pt"),
  s_table_style(background_color = "#CCCCCC", vertical_alignment = "center"))

spec <- add_span_header(spec, cols = c("var1", "var2", "var3"), label = "Baseline",
  stubOrder = 1, labelStyleRef = "level1_stub")

spec <- add_span_header(spec, cols = c("var1", "var2"), label = "Safety",
  stubOrder = 2, labelStyleRef = "level2_stub")
```

---

## Troubleshooting

### Error: "Context error" or "can only be used inside add_style()"

**Problem**: You used an `s_*` helper outside `add_style()`.

```{r error_context, eval = FALSE}
# WRONG
my_style <- s_font(bold = TRUE)  # ERROR

# CORRECT
spec <- add_style(spec, id = "my_style", s_font(bold = TRUE))
```

### Error: "Invalid parameter" or "Allowed values are..."

**Problem**: You used an invalid parameter value (e.g., "bolded" instead of TRUE).

```{r error_invalid, eval = FALSE}
# WRONG
s_font(bold = "bolded")  # ERROR: must be TRUE/FALSE

# CORRECT
s_font(bold = TRUE)
```

### Styles not applied after `create_report()`

**Problem**: If you used `f_combine()`, you must call `create_report()` before the styles are consolidated.

```{r error_consolidate, eval = FALSE}
spec <- create_table(mtcars)
spec <- define_cols(spec, mpg, labelStyleRef = f_combine("bold", "red"))

# WRONG: save directly without consolidation
save_report(spec, ...)  # Combined styles may not be consolidated

# CORRECT: consolidate first
report <- create_report(spec)  # Merges "bold" + "red"
save_report(report, ...)
```

### Styles looking different in renderer than expected

**Problem**: Renderer may override or ignore certain style properties based on its own template system.

**Solution**: Test with simple styles first, then incrementally add properties. Contact your renderer maintainer for constraints.

---

## Advanced notes

### Style consolidation in `create_report()`

When you call `create_report()`, the package:
1. Collects all specs
2. For each spec, finds all style references used with `f_combine()`
3. Merges those combined styles into single consolidated styles
4. Generates unique hash-based names for consolidated styles
5. Updates all references to point to the consolidated style

You don't need to inspect or manipulate `spec$attribs$styles` — the consolidation is automatic and transparent.

### Style reference resolution

Style references are resolved in this order:
1. **Named styles** defined in the same spec with `add_style()`
2. **Predefined clinical styles** (if enabled in options)
3. **Error**: If reference not found, `create_report()` will error with informative message

---

## Built-in style atoms

ksTFL ships a library of small, single-purpose **style atoms** that are available in every
spec without any `add_style()` call.  The design principle is **one property per atom**;
combine atoms with `f_combine()` to build any composite style you need.

### Naming convention

| Prefix | Meaning | Examples |
|---|---|---|
| `b` / `i` / `u` | font bold / italic / underline | `b`, `i`, `u` |
| `fs_N` | font size in pt | `fs_7`, `fs_9`, `fs_11` |
| `fc_<colour>` | font (text) colour | `fc_red`, `fc_navy`, `fc_slate` |
| `hl_<colour>` | text highlight / cell shading | `hl_yellow`, `hl_peach`, `hl_mint` |
| `al` / `ar` / `ac` | paragraph alignment | `al` (left), `ar` (right), `ac` (centre) |
| `ind1` … `ind4` | left-indent levels (0.5 cm steps) | `ind1`, `ind2`, `ind3`, `ind4` |
| `sp_0` / `sp_2` / `sp_4` | spacing before+after in pt | `sp_0`, `sp_2`, `sp_4` |
| `kl` / `kn` | pagination: keep_lines / keep_next | `kl`, `kn` |
| `va_t` / `va_m` / `va_b` | cell vertical alignment | `va_t`, `va_m`, `va_b` |
| `to_h` / `to_90` / `to_270` | cell text orientation | `to_h`, `to_90`, `to_270` |
| `bg_<colour>` | cell background colour | `bg_blue`, `bg_mint`, `bg_steel` |
| `row_h2` / `row_h4` / `row_h6` | row height atoms | separator rows |
| `bt` / `bb` / `bl` / `br` | border side 1 pt black | top / bottom / left / right |
| `bt_th` / `bb_th` | border side 0.5 pt black (thin) | subtotal dividers |
| `bc_gray` / `bc_white` | border colour override | gray / suppress |
| `grp_hdr` / `grp_hdr_i` | group-header composites | bold + space-above |

Long-form aliases are provided for readability:
`font_bold`, `font_italic`, `font_underline`,
`text_left`, `text_right`, `text_center`,
`indent_1` … `indent_4`,
`va_top`, `va_center`, `va_bottom`,
`text_horizontal`, `text_vertical_90`, `text_vertical_270`,
`fc_grey`, `hl_grey`, `bg_grey`, `bc_grey`.

---

### Font decoration

| Style | Property | Typical use |
|---|---|---|
| `b` / `font_bold` | bold | Group headers, totals, primary endpoint values |
| `i` / `font_italic` | italic | Sub-parameter labels, notes |
| `u` / `font_underline` | underline | Hyperlink-style emphasis |

---

### Font size

| Style | Size | Typical use |
|---|---|---|
| `fs_7` | 7 pt | Footnote markers in dense listings |
| `fs_8` | 8 pt | Compact safety tables |
| `fs_9` | 9 pt | Standard clinical table body |
| `fs_10` | 10 pt | Slightly larger body text |
| `fs_11` | 11 pt | Emphasis / primary endpoint result |

---

### Font colour

Pure colours are for strong visual flags; muted/pastel colours suit secondary content
and labels where you want differentiation without distraction.

| Style | Hex | Typical use |
|---|---|---|
| `fc_black` | `#000000` | Reset to default black |
| `fc_red` | `#FF0000` | Out-of-range, critical alert |
| `fc_blue` | `#0000FF` | Informational annotation |
| `fc_green` | `#008000` | Within-range, pass |
| `fc_gray` / `fc_grey` | `#595959` | Secondary values, "(continued)" |
| `fc_navy` | `#1F3864` | Formal headers, dark emphasis |
| `fc_teal` | `#2E75B6` | Softer informational blue |
| `fc_olive` | `#5C7A29` | Muted green — normal range |
| `fc_rust` | `#C0392B` | Muted red — caution (less alarming than pure red) |
| `fc_plum` | `#7B2D8B` | Special category / protocol deviation |
| `fc_slate` | `#44546A` | Subdued blue-gray label |

---

### Text highlight (cell shading)

| Style | Hex | Typical use |
|---|---|---|
| `hl_yellow` | `#FFFF00` | Strong flag / attention |
| `hl_red` | `#FF0000` | Critical value / TEAE |
| `hl_green` | `#90EE90` | Within normal range |
| `hl_gray` / `hl_grey` | `#EEEEEE` | Suppressed / N/A cell |
| `hl_peach` | `#FADADD` | Soft red — mild alert |
| `hl_mint` | `#D5F5E3` | Soft green — normal range |
| `hl_sky` | `#D6EAF8` | Soft blue — informational |
| `hl_lemon` | `#FEFBD8` | Soft yellow — caution |
| `hl_lilac` | `#E8DAEF` | Soft purple — special category |

---

### Paragraph alignment

| Style | Alignment | Alias |
|---|---|---|
| `al` | left | `text_left` |
| `ar` | right | `text_right` |
| `ac` | centre | `text_center` |

---

### Indentation

Each level adds 0.5 cm of left indent — used for stub / sub-group hierarchies.

| Style | Left indent | Alias | Typical use |
|---|---|---|---|
| `ind1` | 0.5 cm | `indent_1` | Top-level category label, e.g. "Age (years)" |
| `ind2` | 1.0 cm | `indent_2` | First sub-group, e.g. "< 18" |
| `ind3` | 1.5 cm | `indent_3` | Second sub-group, e.g. "Missing" |
| `ind4` | 2.0 cm | `indent_4` | Third sub-group / detail level |

---

### Paragraph spacing

| Style | Before | After | Typical use |
|---|---|---|---|
| `sp_0` | 0 pt | 0 pt | Dense safety listings, large tables |
| `sp_2` | 2 pt | 2 pt | Sub-section breathing room |
| `sp_4` | 4 pt | 4 pt | Summary / efficacy tables |

---

### Pagination control

| Style | Property | Typical use |
|---|---|---|
| `kl` | `keep_lines` | Keep all lines of a multi-line cell on the same page |
| `kn` | `keep_next` | Keep this row with the following row (prevents orphaned group headers) |

---

### Group / category header composites

These combine bold + 4 pt space-above + left-indent reset into a single atom.
Equivalent to `f_combine("b", "sp_4")` with an explicit `ind` reset.

| Style | Properties | Typical use |
|---|---|---|
| `grp_hdr` | bold, 4 pt before, 0 pt after, no indent | Category label rows: "Vital Signs", "Age" |
| `grp_hdr_i` | bold + italic, same spacing | Sub-parameter labels within by-visit tables |

---

### Cell vertical alignment

| Style | Alignment | Alias |
|---|---|---|
| `va_t` | top | `va_top` |
| `va_m` | middle (centre) | `va_center` |
| `va_b` | bottom | `va_bottom` |

---

### Cell text orientation

Rotate cell text for narrow column headers in dense visit/parameter grids.

| Style | Rotation | Reads | Alias |
|---|---|---|---|
| `to_h` | 0° (default) | left → right | `text_horizontal` |
| `to_90` | 90° counter-clockwise | bottom → top | `text_vertical_90` |
| `to_270` | 270° counter-clockwise | top → bottom | `text_vertical_270` |

`to_90` is the most common choice for column headers: text ascends from the bottom of
the cell, which matches the reading direction of a rotated page.
`to_270` is useful when you prefer text to descend from the top (e.g. matching a
right-to-left column layout).

```r
# Narrow visit-column header, rotated, centred, bold, bottom-aligned
add_column(spec, "Visit",
           styleRef = f_combine("b", "to_90", "va_b", "ac"))

# Parameter-grid header row — all header cells rotated
header_style <- f_combine("b", "to_90", "va_b", "ac", "bg_steel")
add_column(spec, c("Wk 1", "Wk 2", "Wk 4", "Wk 8"),
           styleRef = header_style)
```

---

### Cell background colour

| Style | Hex | Typical use |
|---|---|---|
| `bg_blue` | `#D9E1F2` | Emphasis rows, section headers |
| `bg_gray` / `bg_grey` | `#F2F2F2` | Alternating rows, sub-totals |
| `bg_peach` | `#FADADD` | Soft red — mild alert row |
| `bg_mint` | `#D5F5E3` | Soft green — normal-range row |
| `bg_sky` | `#D6EAF8` | Soft blue — informational row |
| `bg_lemon` | `#FEFBD8` | Soft yellow — caution row |
| `bg_lilac` | `#E8DAEF` | Soft purple — special category row |
| `bg_navy` | `#1F3864` | Dark navy — strong header band |
| `bg_slate` | `#44546A` | Blue-gray — header band |
| `bg_steel` | `#BDD7EE` | Medium blue — column-group header |

---

### Row height atoms

Used with `add_row()` to create visual separator rows.
Combine with border atoms to control whether a line appears.

| Style | Height | Typical use |
|---|---|---|
| `row_h2` | 2 pt | Near-invisible spacer |
| `row_h4` | 4 pt | Small visual gap |
| `row_h6` | 6 pt | Visible gap between sections |

```r
# Invisible spacer (all borders suppressed)
add_row(spec, "", styleRef = f_combine("row_h2", "bc_white"))

# Thin ruled separator (top border only)
add_row(spec, "", styleRef = f_combine("row_h2", "bb_th"))

# Gray ruled separator
add_row(spec, "", styleRef = f_combine("row_h2", "bb_th", "bc_gray"))
```

---

### Border side atoms

Each atom draws one side at 1 pt black.  Combine multiple sides for a box border.

| Style | Side | Weight |
|---|---|---|
| `bt` | top | 1 pt black |
| `bb` | bottom | 1 pt black |
| `bl` | left | 1 pt black |
| `br` | right | 1 pt black |
| `bt_th` | top | 0.5 pt black (thin) |
| `bb_th` | bottom | 0.5 pt black (thin) |

---

### Border colour override atoms

Apply **after** a side atom to change its colour.  Styles are merged left-to-right
(last-win), so the colour override replaces the colour set by the preceding side atom.

| Style | Colour | Typical use |
|---|---|---|
| `bc_gray` / `bc_grey` | `#AAAAAA` | Softer dividers in banded tables |
| `bc_white` | `#FFFFFF` / none | Suppress / hide all borders (separator rows) |

```r
f_combine("bt", "bc_gray")    # gray 1 pt top border
f_combine("bt", "bb", "bl", "br")          # full black box
f_combine("bt", "bb", "bl", "br", "bc_gray") # full gray box
```

---

### Composition examples

```r
# Sub-group value: indented, italic, muted text
define_cols(spec, "subcat",
  valueStyleRef = f_combine("ind2", "i", "fc_gray"))

# Group header: bold, centered, steel-blue background
define_cols(spec, "category",
  valueStyleRef = f_combine("grp_hdr", "ac", "bg_steel"))

# Out-of-range cell: rust text + peach highlight + bold
c_style(value, styleRef = f_combine("fc_rust", "hl_peach", "b"))

# Subtotal row: thin top border + light gray background + bold
add_row(spec, total_row, styleRef = f_combine("bt_th", "bg_gray", "b"))

# Primary endpoint highlight: full box border + lemon background
c_style(value, styleRef = f_combine("bt", "bb", "bl", "br", "bg_lemon", "b"))

# Keep group header with first data row (no orphan at page bottom)
define_cols(spec, "category", valueStyleRef = f_combine("grp_hdr", "kn"))
```

---

## See also

- Function documentation: `?add_style`, `?s_font`, `?s_paragraph`, `?f_combine`
- [Reporting Examples](Reporting_Examples_with_ksTFL.Rmd) — working examples with styles
- [Getting Started](Getting_Started_with_ksTFL.Rmd) — quick overview

